---
title: "R Cheat Sheets & Notes"
output:
  html_document:
    code_folding: hide
---

## Cheat Sheets

* [R Colors](https://sape.inf.usi.ch/quick-reference/ggplot2/colour)

* [Posit/RStudio Cheat Sheets](https://rstudio.github.io/cheatsheets/html/posit-team.html)

* [R Base Commands Cheat Sheet](https://iqss.github.io/dss-workshops/R/Rintro/base-r-cheat-sheet.pdf)

* [Keyboard Shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts)

* [Quarto Cheat Sheet](https://rstudio.github.io/cheatsheets/html/quarto.html)

* [ggplot2 Cheat Sheet](https://ctl.duke.edu/wp-content/uploads/2020/07/R_ggplot2_cheatsheet.pdf)

* [Math 221 Textbook](https://byuistats.github.io/BYUI_M221_Book/Lesson01.html)

* [Sampling Distribution Applet](https://onlinestatbook.com/stat_sim/sampling_dist/index.html)

* [Normal t Probability Applet](https://byuimath.com/apps/normprobwitht.html?z=-2.601&df=48)

## Test Tips

Drink Gatorade part way through
Have paper and pencil in case you need to rearrange an equation
Read output carefully so you understand what it's saying! 
For anything asking for the correct equation, slow down and double check everything carefully.
For quartemile time (qsec) a lower number is  better.<br>
Read carefully so you don't miss info in the questions.<br>
Make sure to check whether the answer is positive or negative.<br>
Make sure you put a zero before the decimal in Canvas.<br>
Make sure you hit Ctrl+enter
For SSE, etc load them into named variables, so it's easier to work with them ie, SSR/SSTO for R^2
Use the applet to double check t p etc
really look at direction for scientific notation

## Basics
P-value indicates how likely the sample results are assumingthe null hyp is true (It is not the probability the null hyp is correct and is not measure the size or importance of the result)

We use a t distribution when variance is unknown


Type 1: a true null hyp incorrectly rejected
Type 2: a null hypothesis incorrectly accepted
An alpha of .05 means there's a 5% risk of deciding an effect exists when it doesn't
If the P-value is greater than or equal to the critical value, we don't reject the null hypothesis
You can reject the null hypothesis or fail to reject the null hypothesis

Null Hypothesis uppertail for claim saying higher (>), lower tail for claim saying lower/less (<) and two-tailed test for something saying is different ($\neq$)<br>
Test statistic is an number competed from sample used to determine whether to accept/reject the null hypothesis.

## Modules

### 01 Linear Regression
$\beta_0, \beta_1, \sigma^2$ are for the group of dots.
The $i$'s are for the individual.
Since we can't know the errors, we estimated them with the residual.

Three main elements to the mathematical model of regression.

*  The true, unobservable line of the estimated value of the population (regression relation), provides the average Y-value-the expected value of Y:<br> 
$ \underbrace{E\{Y\}}_{\substack{\text{true mean} \\ \text{y-value}}} = \underbrace{\overbrace{\beta_0}^\text{y-intercept} + \overbrace{\beta_1}^\text{slope} X}_\ $<br>

*  The dots (regression plus error term), $Y_i$, $\epsilon_i$ allows each individual $i$ to deviate from the line. $Y_i = \underbrace{\beta_0 + \beta_1 X_i}_{E\{Y_i\}} + \underbrace{\epsilon_i}_\text{error term} \quad \text{where} \ \epsilon_i\sim N(0,\sigma^2)$<br>

*  The estimated line (the line we get from the sample data) 
$\hat{Y}_i$ is the estimated equation and is interpreted as the estimated average $Y$ for any $X$ $\underbrace{\hat{Y}_i}_{\substack{\text{estimated mean} \\ \text{y-value}}} = \underbrace{b_0 + b_1 X_i}_\text{estimated regression equation}$

FORMULA FOR COMPARISON ON TESTS:
$\underbrace{\hat{Y}_i}_{\substack{\text{estimated mean} \\ \text{y-value}}} = b_0 + b_1 \underbrace{X_i}_\text{regression's X value}$

$E\{Y\}$ true mean Y <br>
$Y_i$ the dots (add error term to above), the regression model <br>
$\hat{Y}_i = b_0 + b_1 X_i$, fitted line, lmObject$fitted.values <br>
$b_0$ Estimated intercept, 	b_0 <- mean(Y) - b_1*mean(X) <br>
$b_1$ Estimated slope, b_1 <- sum( X*(Y - mean(Y)) ) / sum( (X - mean(X))^2 ) <br>
$r_i$ residual-eye, distance of dot from line, lmObject's residuals <br>
$\sigma^2$ variance of $\epsilon_i$ <br>

Standard error = estimated standard deviation = the residual standard error in summary
An increase of 1,000 lbs in the weight of a vehicle results in a 5.34 mpg decrease in the average gas mileage of such vehicles.

The interpretation of β1 is the amount of increase (or decrease) in the average y-value, denoted E{Y}, per unit change in X. It is often misunderstood to be the “average change in y” or just “the change in y” but it is more correctly referred to as the “change in the average y”.

In the summary, the sigma is residual standard error and the r2 is Multiple R squared

##### Class Code lm
```{r, eval=FALSE}
# most important for week 1
mylm <- lm(Y ~ x, data=airquality)
summary(mylm)
predict(mylm, data.frame(Wind=19))

# actual class code
library(tidyverse)
View(airquality)
?airquality

# Basics
library(mosaic) #favstats
favstats(airquality$Temp)
mean(airquality$Temp)
sd(airquality$Temp)

# Histograms
hist(airquality$Temp)

ggplot(airquality, aes(x=Temp)) +
geom_histogram(binwidth=5, fill="skyblue", color="skyblue4") +
labs(title="Maximum daily...", subtitle = "May to September 1973", 
     x="Temperature in Degrees F", y="Number of Days in Temperature Range")

# Boxplot 
boxplot(Temp ~ Month, data=airquality)

ggplot(airquality, aes(x=as.factor(Month), y=Temp)) + 
  geom_boxplot()

# Scatterplot
mylm <- lm(Temp ~ Wind, data=airquality)
summary(mylm)
predict(mylm, data.frame(Wind=19))

ggplot(airquality, aes(x=Wind, y=Temp)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F, formula=y~x)

favstats(Temp ~ Month, data=airquality)
ggplot(airquality, aes(x=as.factor(Month), y=Temp)) + 
  geom_boxplot(fill="skyblue", color="skyblue3") + 
  labs(title="Maximum daily temperature in degrees Fahrenheit at La Guardia Airport, NY, USA",
       subtitle="May to September 1973",
       y="Temperature in degrees F", x="Month of the Year")


library(tidyverse)

ggplot(airquality, aes(x=Wind, y=Temp)) + 
  geom_point(fill="skyblue", pch=21, color="skyblue4") + 
  geom_smooth(method="lm", se=F, formula=y~x, color="skyblue4") + 
  labs(title="Max...", subtitle="May...", 
       y="Temp...", x="Average...")

mylm <- lm(Temp ~ Wind, data=airquality)
summary(mylm)






set.seed(101) #Allows us to always get the same "random" sample
#Change to a new number to get a new sample

n <- 153 #set the sample size

X_i <- runif(n, 15, 45) #Gives n random values from a uniform distribution between 15 to 45.

beta0 <- 89 #Our choice for the y-intercept. 

beta1 <- -1.25 #Our choice for the slope. 

sigma <- 1 #Our choice for the std. deviation of the error terms.

epsilon_i <- rnorm(n, 0, sigma) #Gives n random values from a normal distribution with mean = 0, st. dev. = sigma.

Y_i <- beta0 + beta1*X_i + epsilon_i #Create Y using the normal error regression model

fabData <- data.frame(y=Y_i, x=X_i) #Store the data as data

View(fabData) 

#In the real world, we begin with data (like fabData) and try to recover the model that (we assume) was used to created it.

fab.lm <- lm(y ~ x, data=fabData) #Fit an estimated regression model to the fabData.

summary(fab.lm) #Summarize your model. 

plot(y ~ x, data=fabData, ylim=c(20,100)) #Plot the data.

abline(fab.lm) #Add the estimated regression line to your plot.

# Now for something you can't do in real life... but since we created the data...

abline(beta0, beta1, lty=2) #Add the true regression line to your plot using a dashed line (lty=2). 

legend("topleft", legend=c("True Line", "Estimated Line"), lty=c(2,1), bty="n") #Add a legend to your plot specifying which line is which.
```

##### Quizzes lm

```{r, eval=FALSE}
# Assessment Quiz
# 1)
#  library(car), View(Davis), ?Davis
# Reduce the data to just the data for the males. Then perform a regression with weight as the response variable and height as the explanatory variable.
# 
# Which of the following provides the estimated average weight of males that are 180 cm tall?
dav.lm <- lm(weight ~ height, data=subset(Davis, sex=="M"))
predict(dav.lm, data.frame(height=180))

# 2) 
# View(USArrests),  ?USArrests
# Perform a regression using this data that explains the average number of murder arrests in cities (per 100,000 in 1973) using the number of assault arrests (per 100,000) in a city.
#
# Select the answer that provides the correct estimate of B1 in the formula:for the USArrests regression described above.
lm(Murder ~ Assault, data = USArrests) # Choose the assault estimate

#3)
# Which of the following statements is a correct statement about the graphic shown below?
#
# Note: the "Line of Equality" shows where the line would need to be if the reported heights were equal (on average) to the actual measured heights. Dots that fall on this line show men that knew their actual height before they were officially measure
# 3 Answer (note image shows estimated line getting closer to the line of equality as heights increase)
# The average actual height gets closer to the reported height for taller men, while shorter men seem more likely to under-report their height, on average.
```
<br>

### 02 ri...MSE
Lots of notes: 425 Analysis/Residuals.qmd or RCheatSheetsbroken (next to this one)
MSE estimates sigma^2

##### Class Code ri

```{r eval=FALSE}
# most important
ri <- mylm$residuals
sse <- sum(mylm$res^2 ) # SSE is the unexplained variability
ssr <- sum( (mylm$fit - mean(data$Y))^2) # SSR is the explained variablility
ssto <- sum((data$Y - mean(data$Y))^2) # SSTO, variability from ybar
r2 <-ssr/ssto # R2 is the ration betwee explanation and total SSR/SSTO
sqrt(r2) # r is square root of r^2

# how to index dataset
cars[23, ] # row
cars.lm$residuals[23] # corresponding residual
cars$dist[23] - cars.lm$fitted.values[23] #alt to calc single residual

###
# Monday
cars2 <- cars[sample(1:50, 10), ]
View(cars2)

library(tidyverse)

carslm <- lm(dist ~ speed, data=cars2)
summary(carslm)
predict(carslm, data.frame(speed=17.5))

# add line segments
ggplot(cars2, aes(x=speed, y=dist)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F, formula=y~x) + 
  geom_point(aes(x=17.5, y=45.297), color="red", cex=2) + 
  geom_point(aes(x=17.5, y=50), color="blue", cex=2) + 
  geom_segment(aes(x=17.5, xend=17.5, y=45.297, yend=50))


# add squares
ggplot(cars2, aes(x=speed, y=dist)) + 
  geom_point() + 
  geom_smooth(method="lm", se=F, formula=y~x) + 
  geom_point(aes(x=17.5, y=45.297), color="red", cex=2) + 
  geom_point(aes(x=17.5, y=50), color="blue", cex=2) + 
  geom_rect(aes(xmin=17.5, xmax=17.5+.65, ymin=45.297, ymax=50), alpha=0.1) + 
  geom_rect(aes(xmin=speed, xmax=speed+carslm$residuals*.15, ymin=dist, ymax=carslm$fitted.values), alpha=0.1, fill="blue")

install.packages("plotly")
library(plotly)

x = c(5, 15, 2, 29, 35, 24, 25, 39) 
sum(x)
sum( (1:6)^2 ) 
sum(x^2)
mn <- mean(x)
var(x)
dist <-(x-mn)
dist
sum(dist^2)/7
mean(cars$dist)
sum((1:3)^2)

# Wednesday
cars.lm <- lm(dist ~ speed, data=cars)
summary(cars.lm)
cars2 <- cbind(rbind(cars, 
                     cbind(speed=rep(0,50), dist=cars$dist),
                     cbind(speed=cars$speed, dist=cars.lm$fitted.values)),
               frame = rep(c(2,1,3), each=50))
plot_ly(cars2,
        x = ~speed,
        y = ~dist,
        frame = ~frame,
        type = 'scatter',
        mode = 'markers',
        showlegend = F,
        marker=list(color="firebrick")) %>%
  layout(title="Stopping Distance of 1920's Vehicles\n (cars data set)",
         xaxis=list(title="Vehicles Speed in mph (speed)"),
         yaxis=list(title="Stopping Distance in Feet (dist)")) %>%
  add_segments(x = 0, xend = 25, y = mean(cars$dist), yend = mean(cars$dist), line=list(color="gray", dash="dash", width=1), inherit=FALSE, name=TeX("\\bar{Y}"), showlegend=T) %>%
  add_segments(x = 0, xend = 25, y = sum(coef(cars.lm)*c(1,0)), yend = sum(coef(cars.lm)*c(1,25)), line=list(color="darkgray", width=1), inherit=FALSE, name=TeX("\\hat{Y}"), showlegend=T) %>%
  config(mathjax = 'cdn') %>%
  animation_opts(frame=2000, transition=1000, redraw=T)

# Friday
library(tidyverse)
View(diamonds)
plot(price ~ carat, data=diamonds)

diamonds.lm <- lm(price ~ carat, data=diamonds)
summary(diamonds.lm)

abline(diamonds.lm, col="green", lwd=2)

plot(dist ~ speed, data=cars)
cars.lm <- lm(dist ~ speed, data=cars)
summary(cars.lm)
abline(cars.lm)
points(dist ~ speed, data=cars[23,], col="orange", pch=16)
```

##### Quizzes ri


```{r eval=FALSE}
(unsure but this doesn't seem to be the Mobius quiz)
# wt
mtwt.lm <- lm(mpg~wt, mtcars)
summary(mtwt.lm)

ggplot(mtcars, aes(y=mpg, x=wt)) +
  geom_point() +
  geom_smooth(method="lm", se=F, forumula=y~x)

# SSE is the unexplained variability
# sum( lmObject$res^2 )
sum(mtwt.lm$res^2 )

# SSR is the explained variablility
# sum( (lmObject$fit - mean(YourData$Y))^2 )
(mtwt.ssr <- sum( (mtwt.lm$fit - mean(mtcars$mpg))^2 ))

# SSTO
#sum( (YourData$Y - mean(YourData$Y))^2 )
(mtwt.ssto <- sum( (mtcars$mpg - mean(mtcars$mpg))^2 ))


# R2 is the ratio between explanation and total SSR/SSTO
(wtr2 <-mtwt.ssr/mtwt.ssto)


# r is square root of r^2
sqrt(wtr2)

predict(mtwt.lm, data.frame(wt=3*365))

par(mfrow=c(1,3))
plot(mtwt.lm, which = 1:2)
plot(mtwt.lm$residuals)


# cyl
View(mtcars)
mtcyl.lm <- lm(mpg~cyl, mtcars)
summary(mtcyl.lm)
mean(mtcars$mpg)

ggplot(mtcars, aes(y=mpg, x=cyl)) +
  geom_point() +
  geom_smooth(method="lm", se=F, forumula=y~x)

# SSE is the unexplained variability
# sum( lmObject$res^2 )
sum(mtcyl.lm$res^2 )

# SSR is the explained variablility
# sum( (lmObject$fit - mean(YourData$Y))^2 )
(mtcyl.ssr <- sum( (mtcyl.lm$fit - mean(mtcars$mpg))^2 ))

# SSTO
#sum( (YourData$Y - mean(YourData$Y))^2 )
(mtcyl.ssto <- sum( (mtcars$mpg - mean(mtcars$mpg))^2 ))


# R2 is the ration between explanation and total SSR/SSTO
(cylr2 <-mtcyl.ssr/mtcyl.ssto)


# r is square root of r^2
sqrt(cylr2)

mthp.lm <- lm(mpg~hp, mtcars)
summary(mthp.lm)

ggplot(mtcars, aes(y=mpg, x=wt)) +
  geom_point() +
  geom_smooth(method="lm", se=F, forumula=y~x)

# SSE is the unexplained variability
# sum( lmObject$res^2 )
sum(mthp.lm$res^2 )

# SSR is the explained variablility
# sum( (lmObject$fit - mean(YourData$Y))^2 )
(mthp.ssr <- sum( (mthp.lm$fit - mean(mtcars$mpg))^2 ))

# SSTO
#sum( (YourData$Y - mean(YourData$Y))^2 )
(mthp.ssto <- sum( (mtcars$mpg - mean(mtcars$mpg))^2 ))


# R2 is the ratio between explanation and total SSR/SSTO
(wtr2 <-mthp.ssr/mtwt.ssto)


# r is square root of r^2
sqrt(wtr2)

predict(mtcyl.lm, data.frame(cyl=3*365))

par(mfrow=c(1,3))
plot(mtcyl.lm, which = 1:2)
plot(mtcyl.lm$residuals)
```

S1a) 
Type out the mathematical equation for this regression model and label both Y and X in the equation
$$\underbrace{\hat{Y}_i}_\text{circumference} = b_0 + b_1 \underbrace{X_i}_\text{age}$$

```{r eval=FALSE}
# Skills: Residuals, Sums of Squares, and R squared
# 1) Perform a simple linear regression of the circumference of # the tree based on its age in days.
# # mathematical equation is above
View(Orange)
?Orange

#b plot
library(ggplot2)
ggplot(data=Orange, mapping=aes(y=circumference, x=age)) + geom_point(color = "orange", pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()

#c State and interpret the slope and y-intercept of this regression. Are they meaningful for this data?
o.lm <- lm(circumference ~ age, data=Orange)
summary(o.lm)

#d
sum( o.lm$res^2 ) # SSE
sum( (o.lm$fit - mean(Orange$circumference))^2 ) #SSR
sum( (Orange$circumference - mean(Orange$circumference))^2 ) #SSTO
# R2 is Multiple R squared from summary
# correlation is r (sqrt of R2)
osum <- summary(o.lm) 
sqrt(osum$r.squared)
#What circumference would we expect orange trees to have on average after 3 years of growth (assume 365 days = 1 year) based on this regression?
# note: age is in days so 3 X 365
predict(o.lm, data.frame(age=3*365))

#just in case
## sqrt(MSE)
sqrt(sum(o.lm$residuals^2)/(35-2))
plot(o.lm, which=1:2)
plot(o.lm$residuals)
# the variance is not constant, minor issues with linearity, minor issues with normality

# S1e Both of these are true.
# The SSR divided by the SSTO gives the proportion of variation in Y explained by the regression, which is R-squared. The square root of R-squared is the correlation.
# The square root of SSR divided by SSTO gives the value of the correlation.
```


```{r eval=FALSE}

# Assessment: Residuals, Sums of Squares, and R squared
# 1) A regression was performed for a sample of n = 5 data points.
# The y-values of the regression are: 3.78, 6.08, 6.65, 9.25, and 9.92.
# The residuals from the regression are: -0.266, 0.489, -0.486, 0.569, and -0.306.
# What is the R-squared value for this regression?
y <- c(3.78, 6.08, 6.65, 9.25, 9.92)

SSTO <- sum( (y - mean(y))^2 )
#SSTO = 24.83372

res <- c(-0.266, 0.489, -0.486, 0.569, -0.306)
SSE <- sum(res^2)
#SSE = 0.96347

R2 = 1 - SSE/SSTO
#R2 = 0.9612032

# 2) This data can be used to show that the displacement of the engine (disp)
# is positively correlated with the weight of the vehicle (wt).
# Apparently, this means disp is the Y and wt is the x
# Don't use mylm unless you've cleared everything or use something descriptive like clm for cars mylm
mt.lm <- lm(disp ~ wt, data=mtcars)
summary(mt.lm)
# Multiple R-squared: 0.7885, Adjusted R-squared: 0.7815  <- (multiple R-squared is R^2)

# 3)
# A certain statistics teacher at BYU-Idaho drives a 2001 Nissan Sentra that weighs
# approximately 2,700 lbs and currently gets only 21 mpg.
# Based on the regression you performed, how many mpg above or below average is this vehicle?
#   Perform a regression of mpg means mpg is the Y and wt is the x
mt.lm <- lm(mpg ~ wt, data=mtcars) #perform the regression
plot(mpg ~ wt, data=mtcars) #draw the regression (not needed, but nice)
abline(mt.lm) #add the regression line (not needed, but nice)
points(2.7, 21, pch=16, col="skyblue") #add the Nissan Sentra value (not needed, but nice)
predict(mt.lm, data.frame(wt=2.7)) #get predicted value for Nissan Sentra
1
22.85505
lines(c(2.7, 2.7), c(21, 22.85505), col="skyblue") #add residual line (not needed, but nice)
myresidual <- 21 - 22.85505 #calculate difference between Y and Y-hat

> myresidual
[1] -1.85505
```
<br>



### 03 Diagnostics & Transformations

##### Residual Plots & Regression Assumptions & Transformations

*  residuals vs. fitted-values (which=1), most important (for assump 1 straight line,3 no pattern)
*  Q-Q Plot of Residuals (which=2), least important (for assump 2, close to line) or library(car), qqPlot(lm$residuals) has guides
*  Residuals vs Order (lm$residals), second most important (for assump 5, no pattern)<br>


Five Assumptions:

1)  The regression relation between Y and X is linear, plot(mylm, which=1)<br>
*res v fit (red line straight) If bent results are meaningless--it affects everything.*<br>
Nothing can be trusted if the linearity assumption is violated.
2)  The error terms are normally distributed with E{ϵi}=0. QQ checks normality<br>
plot(mylm, which=2)<br>
*dots should be close to on the line, watch tails*<br>
If the error terms are not normal, the slope and intercept are often still meaningful (when the independence assumption is violated), but the residual standard error could be unnecessarily large in this case. 
3)  The variance of the error terms is constant over all X values, plot(mylm, which=1)<br>
*no pattern in plot?*<br>
Distance should be constant on res v fit. When variance of the error term changes across the regression, the regression approximates the “average variance” of the errors because the regression is still assuming the variance is constant across the regression. The estimates of the slope and intercept are still typically quite good, and can be used for interpretation. The residual standard error however should not be considered to be meaningful as it will be too large on one end of the regression and too small on the other end.<br> 
4)  The X values can be considered fixed and measured without error.
5)  The error terms are independent. plot(mylm$residuals, ylab="Residuals")<br>
*res v order should have no pattern*<br>
While the slope and intercept are often still meaningful when the independence assumption is violated, the residual standard error is unnecessarily large in this case.Can be checked when the data has natural ordering to it eg., time series often break it.<br>

Also--WATCH FOR OUTLIERS, While outliers do not violate any of the regression assumptions, they do pose substantial difficulties for the least squares regression estimates of the slope and intercept.

$$Y_i = \underbrace{\beta_0 + \beta_1 \overbrace{X_i}^\text{#4}}_{\text{#1}} + \epsilon_i \quad \text{where} \ \overbrace{\epsilon_i \sim}^\text{#5} \overbrace{N(0}^\text{#2}, \overbrace{\sigma^2}^\text{#3})$$


How does plot change when we look at residuals vs fitted-values Which is the y
How check assumption 4 Does it make sense th.n having x fixed is workes
Explain the difficulties that arise when there is an outlier present. (how do outliers skew) biases the slope

##### Transformations
For week 03 assign, tell what BoxCox says, and whether you used it.
Show math, my y` is this, my belief on the prime space is this, resulting equation back in original units is this.

the log changes the way I see the data, changes the lens, changes my perspective. 
Viewing a histogram of the islands data and log islands data--you can see more detail with the log view

Y transformations can be found using boxCox
boxCox(lm)     # from library(car), log-likelihood
boxCox is a suggestion not a rule or law
If no y transform needed, bCox will be around 1

X transformations are more difficult to find because boxCox doesn't work. If linearity is violated, but variance across residuals is constant (both are on rvf which=1)

When have a choice choose log transform because it's the only one that's interpretable. Also choose wisely, for orange trees y^2 trees have no circumference until day 200, 1/y don't really start growing for years and then take over:) For youngOrange, even though log was best for current data it may make less sense for a real set of random trees, y might be the best of these options. Make sure it's logical.

He used the ` prime mark for optimus prime ie. it's been transformed. So Y`i = b0+b1 from other space
Run a lm of transform Y, then summarize : summary(lm.t)
save coef

Interesting note from gpt
When you back transform, this is actually estimating the median of the expression, not the mean, unless you apply a bias correction (interesting technicality)
Mean Squared Error
$$\begin{equation}
  s^2 = MSE = \frac{SSE}{n-2} = \frac{\sum(Y_i-\widehat{Y}_i)^2}{n-2} = \frac{\sum r_i^2}{n-2}
\end{equation}$$

##### Class Code Diag

```{r, eval=FALSE}
# most important code
par(mfrow=c(1,3))
plot(mylm, which = 1:2) #1,3 & 2
plot(mylm$residuals) #5
boxCox(lm) # lambda for Y

###
# Beg week- Diagnosing Models
view(mtcars)
?mtcars
lm.mt <- lm(mpg ~ qsec, data=mtcars)
plot(mpg ~ qsec, data=mtcars)
abline(lm.mt)
plot(lm.mt, which=1)

plot(drat ~ wt, data=mtcars)
lm2 <- lm(drat ~ wt, data=mtcars)
abline(lm2)
plot(lm2, which=1)

?Loblolly
view(Loblolly)
plot(height ~ age, data=Loblolly)
lmlob <- lm(height ~ age, data=Loblolly)
abline(lmlob)
plot(lmlob, which=1)
par(mfrow=c(1,3))
plot(lmlob, which = 1:2)
plot(lmlob$residuals)

summary(lmlob)

?Orange
View(Orange)
plot(circumference ~ age, data=Orange)

#Midweek- Intro to Transformations (log)
#remember log default is natural log, so log(5000), is e^?=5000
#exp(number), is e^number=what
log(5000)
exp(8.517193)
exp(1) #2.718282

hist(islands, col="forestgreen")
hist(log(islands), col="forestgreen")
exp(4)
log(10000)

# End of week
# Graphing Transformations
library(mosaicData)
library(car)
View(Utilities)
?Utilities
# data with lm
gas.lm <- lm(gasbill ~ temp, data=Utilities)
plot(gasbill ~ temp, data=Utilities)
abline(gas.lm)
# res v fitted
plot(gas.lm, which=1) # bend in rvf, he says the outliers have notes, could be removed as bad data
#NEW
boxCox(gas.lm) # to find lambda for Y
lambda <- -2
plot(gasbill^lambda ~ temp, data=Utilities)

# test multiple lambda's to see which is best
for (lambda in c(-2, -1, -.5, -.4, -.3, -.2, -.1, .1, .25, .5, 1, 2)) {
  plot(gasbill^lambda ~ temp, data=Utilities, main = paste("lambda = ", lambda))
}


for (lambda in seq(0.5, 0.005, length.out=10)) {
  plot(gasbill^lambda ~ temp, data=Utilities, main = paste("lambda = ", lambda))
}

plot(log(gasbill) ~ temp, data=Utilities)
u.lm.t <- lm(log(gasbill) ~ temp, data=Utilities)
abline(u.lm.t)

summary(u.lm)$coef
summary(u.lm.t)$coef

exp(predict(u.lm.t, data.frame(temp=90)))
predict(u.lm, data.frame(temp=90))


plot(gasbill ~ temp, data=Utilities)
curve(exp(6 - 0.04*x), add=TRUE, col="skyblue")


ggplot(Utilities, aes(x=temp, y=gasbill)) + 
  geom_point() + 
  stat_function(fun=function(x) exp(6 - 0.04*x))

# OLDdata with log transformation (sqrt, sqrt had best lambda)
gas.lm.log <- lm(log(gasbill) ~ temp, data=Utilities)
b <- coef(gas.lm.log) # grab line
# untransform add line
plot(log(gasbill) ~ temp, data=Utilities)
abline(gas.lm.log)

plot(gas.lm.log, which=1)
summary(gas.lm.log)

#bring back the curved line
plot(gasbill ~ temp, data=Utilities) # base plot, to log
curve(exp(6.031885 - 0.041435*x), add=TRUE) # curve from log, but exp

plot(gasbill ~ temp, data=Utilities)
abline(gas.lm)
curve(exp(b[1] + b[2]*x), add=TRUE, col="skyblue") #curve() requires lower-case x


ggplot(Orange, aes(x=age, y=circumference)) + 
  geom_point(color="orangered") +
  labs(title="Growth of Orange Trees", x="Age of Tree in Days", y="Circumference of Tree (mm)") + 
  theme_bw( )

  stat_function(fun=function(x) ())# not sure what this was for

# Bro Saunders code
lm.log <- lm(log(circumference) ~ age, data=Orange)
b.log <- coef(lm.log)
b.log[1]
b.log[2]

lm.sqrt <- lm(sqrt(circumference) ~ age, data=Orange)
b.sqrt <- coef(lm.sqrt)

lm.1oY <- lm(1/circumference ~ age, data=Orange)
b.1oY <- coef(lm.1oY)

lm.y <- lm(circumference ~ age, data=Orange)
b.y <- coef(lm.y)

lm.y2 <- lm(circumference^2 ~ age, data=Orange)
b.y2 <- coef(lm.y2)

lm.1oY2 <- lm(1/circumference^2 ~ age, data=Orange)
b.1oY2 <- coef(lm.1oY2)


# Base Graphic


plot(circumference ~ age, data=Orange, pch=16, col="orangered", main="Growth of Orange Trees", xlab="Age of Tree in Days", ylab="Circumference of Tree (mm)")

curve( exp(b.log[1] + b.log[2]*x), add=TRUE, col="red")
curve( (b.sqrt[1] + b.sqrt[2]*x)^2, add=TRUE, col="blue")
curve( 1/(b.1oY[1] + b.1oY[2]*x), add=TRUE, col="green")
curve( b.y[1] + b.y[2]*x, add=TRUE, col="gray")
curve( sqrt(b.y2[1] + b.y2[2]*x), add=TRUE, col="orange")
curve( 1/sqrt(b.1oY2[1] + b.1oY2[2]*x), add=TRUE, col="forestgreen")

# ggplot Graphic

library(tidyverse)

ggplot(Orange, aes(x=age, y=circumference)) + 
  geom_point(color="orangered") +
  stat_function(fun=function(x) exp(b.log[1] + b.log[2]*x), aes(color="log(Y)")) + 
  stat_function(fun=function(x) (b.sqrt[1] + b.sqrt[2]*x)^2, aes(color="sqrt(Y)")) + 
  stat_function(fun=function(x) 1/(b.1oY[1] + b.1oY[2]*x), aes(color="1/Y")) + 
  stat_function(fun=function(x) b.y[1] + b.y[2]*x, aes(color="Y")) + 
  stat_function(fun=function(x) sqrt(b.y2[1] + b.y2[2]*x), aes(color="Y^2")) + 
  stat_function(fun=function(x) 1/sqrt(b.1oY2[1] + b.1oY2[2]*x), aes(color="1/Y^2")) + 
  labs(title="Growth of Orange Trees", x="Age of Tree in Days", y="Circumference of Tree (mm)") + 
  theme_bw( ) 

library(car)
boxCox(lm.y)


YoungOrange <- filter(Orange, age < 1200)


lm.log <- lm(log(circumference) ~ age, data=YoungOrange)
b.log <- coef(lm.log)

lm.sqrt <- lm(sqrt(circumference) ~ age, data=YoungOrange)
b.sqrt <- coef(lm.sqrt)

lm.1oY <- lm(1/circumference ~ age, data=YoungOrange)
b.1oY <- coef(lm.1oY)

lm.y <- lm(circumference ~ age, data=YoungOrange)
b.y <- coef(lm.y)

lm.y2 <- lm(circumference^2 ~ age, data=YoungOrange)
b.y2 <- coef(lm.y2)

lm.1oY2 <- lm(1/circumference^2 ~ age, data=YoungOrange)
b.1oY2 <- coef(lm.1oY2)



ggplot(Orange, aes(x=age, y=circumference)) + 
  geom_point(color="orangered") +
  geom_vline(aes(xintercept=1200)) + 
  stat_function(fun=function(x) exp(b.log[1] + b.log[2]*x), aes(color="log(Y)")) + 
  stat_function(fun=function(x) (b.sqrt[1] + b.sqrt[2]*x)^2, aes(color="sqrt(Y)")) + 
#  stat_function(fun=function(x) 1/(b.1oY[1] + b.1oY[2]*x), aes(color="1/Y")) + 
  stat_function(fun=function(x) b.y[1] + b.y[2]*x, aes(color="Y")) + 
  stat_function(fun=function(x) sqrt(b.y2[1] + b.y2[2]*x), aes(color="Y^2")) + 
#  stat_function(fun=function(x) 1/sqrt(b.1oY2[1] + b.1oY2[2]*x), aes(color="1/Y^2")) + 
  ylim(c(0,300)) + 
  labs(title="Growth of Orange Trees", x="Age of Tree in Days", y="Circumference of Tree (mm)") + 
  theme_bw( ) 

# 1. Fit the model with simultaneous transformations
# lambda = 0.25 for y, log for x
model <- lm(I(y^0.25) ~ log(x), data = your_data)
```


##### Quizzes Diag
Skills Quiz


Problem 1

Open the `Davis` dataset in R, found in `library(car)`. As stated in the help file for this data set, "The subjects were men and women engaged in regular exercise." 

Perform a simple linear regression of the height of the individual based on their weight.

1(a) {done}
Type out the mathematical equation for this regression model and label both $Y$ and $X$ in the equation.
$$
\underbrace{\hat{Y}_i}_\text{height} = b_0 + b_1 \underbrace{X_i}_\text{weight}
$$

1(b) {done}
```{r, eval=FALSE}
#Plot a scatterplot of the data with your regression line overlaid.
library(car)
ggplot(data=Davis, mapping=aes(y=height, x=weight)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()

```
1(c) {done}
Create a residuals vs fitted-values plot for this regression. What does this plot show? Ans B, The point in the bottom right corner of the plot labeled "12" (observation 12) appears to be a dramatic outlier causing the regression line to be pulled towards it.
```{r, eval=FALSE, fig.height=4, fig.width=4}
d.lm <- lm(height ~ weight, data=Davis)
summary(d.lm)
plot(d.lm, which=1)

```
1(d) {done}
State and interpret the slope, y-intercept, and $R^2$ of this regression. Are they meaningful for this data under the current regression?
```{r}
#slope 0.15086 is the increase in the average height
#intercept 160.09312  The y-intercept is the average height of exercising individuals who have a weight of zero. In other words, it is not interesting in this analysis.
# R2 0.03597   The proportion of variation in height explained by this regression (with the outlier included) is essentially 0  
# Are these values meaningful for this data under the current regression? 	No, these values are not currently very meaningful because they are being strongly skewed by the presence of the outlier (observation number 12).

```


1(e) {}
Run `View(Davis)` in your Console. What do you notice about observation #12 in this data set? 

Perform a second regression for this data with observation #12 removed. Recreate the scatterplot of Part (b) with two regression lines showing this time. The first regression line should include the outlier. The second should exclude the outlier. Include a legend to show which line is which.
```{r}
View(Davis)
Davis2 <- Davis[-12,]

ggplot(data=Davis, aes(x=weight, y=height)) +
  geom_point(color="gray", pch=16) +
  geom_smooth(aes(color="Regression with outlier"),
              method="lm", se=F) +
  geom_smooth(data=Davis2, 
              aes(x=weight, y=height, color="Regression without outlier"),
              method="lm", se=F) +
scale_color_manual(values = c("Regression with outlier"="grey",
                              "Regression without outlier"="skyblue")) +
  labs(color = "Model") +
  theme_minimal()
```

1(f) {}
Compute the slope, y-intercept, and $R^2$ value for the regression with the outlier removed. compare the results to the values when the outlier was present.
```{r}
d2.lm <- lm(height ~ weight, data=Davis2)
summary(d2.lm)


#slope 0.51689
#intercept 136.83661
#R2 0.594
```
Part (g) ASK BRO SAUNDERS why A not B

Create a residuals vs fitted-values plot for the regression with the outlier removed. How do things look now? Ans This plot shows possible slight problems with linearity (a slight, but consistent bend in the red line) but there do not appear to be any problems with the variance or with outliers. (not sure why not Everything looks really good now, although point #21 is now another outlier that should probably be removed from the regression.)


```{r}
plot(d2.lm, which=1)

```
Problem 2 {}

Open the **Prestige** data set found in `library(car)`.

Perform a regression that explains the 1971 average annual **income** from jobs according to their "Pineo-Porter **prestige** score for occupation, from a social survey conducted in the mod-1960's."

2(a) {}
Plot the data and fitted simple linear regression line.
```{r}
library(car)
View(Prestige)
?Prestige
p.lm <- lm(income~prestige, data=Prestige)
ggplot(data=Prestige, mapping=aes(y=income, x=prestige)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()
```

2(b) {}
State the estimated values for $\beta_0$, $\beta_1$, and $\sigma$ for this regression. Note, the residual standard error is the sigma

```{r}
summary(p.lm)
#B0 -1465.03 
#B1 176.43
#sigma 2984
```

2(c) {}
Create a residuals vs fitted-values plot and a Q-Q Plot of the residuals for this regression. Ans 	
Normality of the error terms is violated, as shown by the points going "out of bounds" in the Q-Q Plot. However, these problems are likely due to the increasing variance of the residuals shown in the residuals vs fitted-values plot. There even may be some problems with linearity of the data because of the three outliers in the top right of the graph pulling the regression line up (shown by the red line going down in Residuals v Fitted values).
The effect: 
	
The slope and estimate of the error variance are almost certainly being negatively effected by the increasing variance and three outliers.

```{r, fig.width=4, fig.height=4}
plot(p.lm, which=1:2, add.smooth=TRUE) #did not work
qqPlot(p.lm) #did work
```

2(d) {}
Comment on any difficulties the diagnostic plots in Part (c) reveal about the regression. 
Comment on which estimates of Part (b) are likely effected by these difficulties.



Problem 3 {}
Open the **Burt** data set from library(car).

This data set is famous for being fraudulent, or fake. See ?Burt for more details. One of the first indicators that it was fraudulent was revealed by regressing IQbio ~ IQfoster. This regression was just a little too good to be real. (Note that for social science data, like this data, $R^2$ values above 0.3 are impressive. Values above 0.7 are rare.)

3(a) {}
Plot the data and fitted regression line. State the estimated values of $\beta_0$, $\beta_1$, and $\sigma$ as well as the $R^2$ of the regression.
```{r}
library(car)
View(Burt)
?Burt
burt.lm <- lm(IQbio~IQfoster, data=Burt)
ggplot(data=Burt, mapping=aes(y=IQbio, x=IQfoster)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()
```

3(b) {}
Create a (1) residuals vs. fitted-values plot, (2) Q-Q Plot of the residuals, and (3) residuals vs. order plot for this regression. Are any problems with regression assumption violations visible in these plots?
```{r, fig.height=4, fig.width=4}
plot(burt.lm, which = 1:2) #1,3 & 2
plot(burt.lm$residuals) #5
```

3(c) {}
Comment on what the three diagnostic plots of Part (b) show for the regression. 
	
They show a fairly nice regression. There may be some slight difficulties with linearity due to the curved red line in the residuals vs fitted-values plot, and points 23 and 24 are somewhat far away from the rest of the data, but otherwise, things are impressively nice.



Problem 4
Open the **mtcars** data set in R.
Perform a regression of **mpg** explained by the **disp**lacement of the vehicle's engine.

4(a) {}
Plot the data and fitted regression line. State the estimated values of $\beta_0$, $\beta_1$, and $\sigma$ as well as the $R^2$ of the regression.
```{r}
View(mtcars)
?mtcars
m.lm <- lm(mpg~disp, data=mtcars)
ggplot(data=mtcars, mapping=aes(y=mpg, x=disp)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()
summary(m.lm)
#B0 29.599855
#B1 -0.041215
#RSE/sigma 3.251
#R2 0.7183
```

4(b) {}
Create a (1) residuals vs. fitted-values plot, (2) Q-Q Plot of the residuals, and (3) residuals vs. order plot for this regression. Are any problems with regression assumption violations visible in these plots?

```{r, fig.height=4, fig.width=4}
plot(m.lm, which = 1:2) #1,3 & 2
plot(m.lm$residuals) #5
```

4(c) {}
Comment on what the three diagnostic plots of Part (b) show for the validity of the values computed in Part (a). 

Linearity doesn't look good. Variance doesn't look consistant.The other two plots look good to me.


Problem 5 {}
Open the **Orange** data set found in R.
Perform a regression that explains the **circumference** of the trunk of the orange tree as the tree **age**s.

4(a) {}
Plot the data and fitted simple linear regression line.
```{r}
View(Orange)
?Orange
o.lm <- lm(circumference~age, data=Orange)
summary(o.lm)
ggplot(data=Orange, mapping=aes(y=circumference, x=age)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()

#B0 17.399650
#B1 0.106770
#sqrt(MSE) or RSE 
```

4(b) {}
State the estimated values for $\beta_0$, $\beta_1$, and $\sigma$ for this regression. 
```{r, fig.width=4, fig.height=4}

```

4(c) {}
Create a residuals vs fitted-values plot and a Q-Q Plot of the residuals for this regression. 
```{r, fig.width=4, fig.height=4}
plot(o.lm, which = 1:2) #1,3 & 2
plot(o.lm$residuals) #5
```

4(d) {}
Comment on any difficulties the diagnostic plots in Part (c) reveal about the regression. 

Comment on which estimates of Part (b) are likely effected by these difficulties.

linearity only slightly off, but no constant variance, qqplot okay, not so sure about the last one, almost looks like lines.

4(e) {}
Perform a Box-Cox analysis of the regression. Which Y-transformation is suggested?

```{r}
# lambda = .5 (sqrt)
boxCox(o.lm)
```

Type your answer here...

</div> 

 
4(f) {}

Perform a regression with the transformed y-variable. Plot the regression in the transformed units. Diagnose the fit of the regression on the transformed data.

Ans Linearity is more violated in the transformed regression than it was in the original units. Normality is more violated in the transformed regression than it was in the original units.

```{r, fig.height=4,fig.width=4}
#lm(sqrt(Y) ~ X)
o.lm.sqrt <- lm(sqrt(circumference)~age, data=Orange)
summary(o.lm.sqrt)
b <- coef(o.lm.sqrt)

ggplot(data=Orange, mapping=aes(y=circumference, x=age)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
stat_function(fun=function(x) (b[1] + b[2]*x)^2, aes(color="sqrt(Y)")) +
  theme_minimal()

plot(o.lm.sqrt, which = 1:2) #1,3 & 2
plot(o.lm.sqrt$residuals)
```




4(g) {}

Write out the fitted model for $\hat{Y}_i'$. Then solve the transformed model back into the original units for $\hat{Y}_i$. Then compute the following.

When $X_i = 500$, then $\hat{Y}_i = \ldots$.


$$
  \hat{Y}_i' = 5.3408 + 0.0055
$$ 

rounded to 4 places


Extra possible from 2023
``` {r, eval=FALSE}
View(Orange)
o.lm <- lm(circumference ~ age, data=Orange)
summary(o.lm)

# Y TRANSFORM IN ORIGINAL SCALE
ggplot(data=Orange, mapping=aes(y=circumference, x=age)) + geom_point(color = "orange", pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
stat_function(fun=function(x) (b[1] + b.[2]*x)^2, aes(color="sqrt(Y)")) + 
  theme_minimal()

# sqrt(MSE)
sqrt(sum(o.lm$residuals^2)/(35-2))

plot(o.lm, which=1:2)
plot(o.lm$residuals)
# the variance is not constant, minor issues with linearity, minor issues with normality

library(car)
boxCox(o.lm)
o.sqrt.lm <- lm(sqrt(circumference) ~ age, data=Orange)
summary(o.sqrt.lm)
b <- coef(o.sqrt.lm)
t.lm <- lm((b[0] + b[1]*age)^2 ~ age, data=Orange)

ggplot(data=Orange, mapping=aes(y=sqrt(circumference), x=age)) + geom_point(color = "orange", pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()

plot(o.sqrt.lm, which=1:2)
plot(o.sqrt.lm$residuals)

plot(circumference ~ age, data=Orange)
abline(o.lm)
curve((b[1] + b[2]*x)^2, add=TRUE)

ggplot(data=Orange, mapping=aes(y=circumference, x=age)) + geom_point(color = "orange", pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  
  theme_minimal()

o.pred <- predict(o.sqrt.lm, data.frame(age=500))
o.pred^2  # use predict on the transformed model and then reverse the answer
```

<br>

### 04 Hypothesis Test for Model Parameters
Sampling Distributions of Model Parameters
Important to understand: ALL STATISTICS ARE WRONG! (because we have a sample)

2 biggest problems in statistics:
1) Every sample is wrong (but they are close)
2) No 2 samples agree  
When you take a sample from the population the line varies from the truth.
The regression lines from samples are closer in the middle and further apart on the ends.

Residual Standard Error is the average spread of the dots from the line (sigma)
Multiple R-squared is the proportion of variability in Y that's explained by the regression
The Estimated Intercept, x, RSE, and R^2 tells us about our data.e The Std. Error, tvalue, and Pr (p), tell us about all possible samples around the truth. "One sample has enough DNA to recreate the whole dinosaur."

repeating (rep) a sequence (seq) is the best way to collect data for regression (create samples), you get two replicates for each place--gives the most insight for the least data possible

It can be mathematically proven the the expected value of b1 is beta1

Only 3 things affect the final variability of the slope and the intercept, sigma (not in our control), sample size, and the spread of x

MSE is the avg standard error and also a measure of the vertical variability of the dots around the line? y?

${\sigma_{b_1}}^2 = \frac{\sigma^2}{sum(X_i-\bar{X})^2}$ The variability of the slope is the variability of the y's over the variability of the x's
MSE is an approximation of sigma squared

Pander is in the pander package
pander(summary(cars.lm)$coefficients) produces a table which has the info necessary to calculate t and p values, see Inference for the Model Parameters in the Statistic Notebook

For hypothesis questions see Inference for Model Parameters remember pt() only calculates 1 tail, so it has to be doubled if our alt hypothesis is not equal

##### Class Code Samp Dist

```{r, eval=FALSE}
### Week Start
# inline r chunk `r beta_0` can use it to populate the equation
# Sampling Distributions of Model Parameters
#for loop
N <- 512 
storage <- rep(NA, N)
storage

for (i in 1:N){
  storage[i] <- 2*i
  cat("i =", i, " and 2*i =", 2*i, " was saved in storage[", i, "]\n")
}

storage

# abline stands for add b-slope line
# geom_smooth, se=F, takes the band away, this week remove, so we can see the band the band suggests where the true line will be


library(tidyverse)

set.seed(121)


N <- 50000
beta0 <- 3
beta1 <- 2.5

sigma <- 1.2

X <- rnorm(N, 30, 5)
Y <- beta0 + beta1*X + rnorm(N, 0, sigma)

plot(Y ~ X, col=rgb(.1,.1,.1,.01), pch=16)

n <- 5
mysample1 <- sample(N, n)

points(Y[mysample1] ~ X[mysample1], col="hotpink", pch=16)
mylm1 <- lm(Y[mysample1] ~ X[mysample1])
summary(mylm1)
abline(mylm1, col="hotpink")
abline(beta0, beta1, col="green1")


n <- 5
mysample2 <- sample(N, n)

points(Y[mysample2] ~ X[mysample2], col="lemonchiffon", pch=16)
mylm2 <- lm(Y[mysample2] ~ X[mysample2])
summary(mylm2)
abline(mylm2, col="lemonchiffon")

mydata <- data.frame(Y=Y[mysample2],X=X[mysample2])

ggplot(mydata, aes(x=X, y=Y)) + 
  geom_point() + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_abline(aes(intercept=beta0, slope=beta1), color="green1")


N <- 512  
storage <- rep(NA, N)
storage

for (i in 1:N){
  storage[i] <- 2*i
}

storage


N <- 5000
storage_int <- storage_slope <- rep(NA, N)

for (i in 1:N){

  #Hint 1

  n <- 40
  Xi <- rep(seq(30, 100, length.out=n/2), each=2) #n must be even.
  Yi <- 2.5 + 3*Xi + rnorm(n, 0, 1.2)

  #Hint 2

  mylm <- lm(Yi ~ Xi)
  #coef(mylm)
  storage_int[i] <- coef(mylm)[1] #intercept only
  storage_slope[i] <- coef(mylm)[2] #slope only

}

#Hint 3
hist(storage_int)
hist(storage_slope)

mean(storage_int)
mean(storage_slope)
sd(storage_int)
sd(storage_slope)

# If I increase the spread of x in the regression model, the sd decreases. 
# If I decrease the spread of x in the regression model, the sd increases.
# If I increase the sigma, the sd increases.
# If I decrease the sigma, the sd decreases.
# reducing samle size increases variability
```

##### Class Code Hyp Test
```{r, eval=FALSE, fig.height=8, fig.width=8}

```{r}
## Simulation to Show relationship between Standard Errors

##-----------------------------------------------
## Edit anything in this area... 

n <- 100 #sample size (choose a decent size sample)
# spread of x (choose a decent spread)
Xstart <- 30 #lower-bound for x-axis 
Xstop <- 100 #upper-bound for x-axis

beta_0 <- 2 #choice of true y-intercept
beta_1 <- 3.5 #choice of true slope
sigma <- 13.8 #choice of st. deviation of error terms (choose a relatively small sigma)

## End of Editable area.

X <- rep(seq(Xstart,Xstop, length.out=n/2), each=2) #Create X - this way most insight for least data 
N <- 5000 #number of times to pull a random sample (for the assignment it's 3)
# notice using b0, b1 because it's not the law, not beta
storage_b0 <- storage_b1 <- rep(NA, N)
for (i in 1:N){
  Y <- beta_0 + beta_1*X + rnorm(n, 0, sigma) #Sample Y from true model
  mylm <- lm(Y ~ X)
  storage_b0[i] <- coef(mylm)[1]
  storage_b1[i] <- coef(mylm)[2]
}

# only works if you're using base r
layout(matrix(c(1,1,2,3), 2, 2, byrow = TRUE), widths=c(2,2), heights=c(3,3))

Ystart <- 0 #min(0,min(Y)) 
Ystop <- 500 #max(max(Y), 0)
Yrange <- Ystop - Ystart
# need plot and histograms, the rest is extra
plot(Y ~ X, xlim=c(min(0,Xstart-2), max(0,Xstop+2)), 
     ylim=c(Ystart, Ystop), pch=16, col="gray",
     main="Regression Lines from many Samples\n Plus Residual Standard Deviation Lines")
text(Xstart, Ystop, bquote(sigma == .(sigma)), pos=1)
text(Xstart, Ystop-.1*Yrange, bquote(sum ((x[i]-bar(x))^2, i==1, n) == .(var(X)*(n-1))), pos=1)
text(Xstart, Ystop-.25*Yrange, bquote(sqrt(MSE) == .(mean(storage_rmse))), pos=1)


for (i in 1:N){
  abline(storage_b0[i], storage_b1[i], col="darkgray")  
}
abline(beta_0, beta_1, col="green", lwd=3)
abline(beta_0+sigma, beta_1, col="green", lwd=2)
abline(beta_0-sigma, beta_1, col="green", lwd=2)
abline(beta_0+2*sigma, beta_1, col="green", lwd=1)
abline(beta_0-2*sigma, beta_1, col="green", lwd=1)
abline(beta_0+3*sigma, beta_1, col="green", lwd=.5)
abline(beta_0-3*sigma, beta_1, col="green", lwd=.5)

par(mai=c(1,.6,.5,.01))

  addnorm <- function(m,s, col="firebrick"){
    curve(dnorm(x, m, s), add=TRUE, col=col, lwd=2)
    lines(c(m,m), c(0, dnorm(m,m,s)), lwd=2, col=col)
    lines(rep(m-s,2), c(0, dnorm(m-s, m, s)), lwd=2, col=col)
    lines(rep(m-2*s,2), c(0, dnorm(m-2*s, m, s)), lwd=2, col=col)
    lines(rep(m-3*s,2), c(0, dnorm(m-3*s, m, s)), lwd=2, col=col)
    lines(rep(m+s,2), c(0, dnorm(m+s, m, s)), lwd=2, col=col)
    lines(rep(m+2*s,2), c(0, dnorm(m+2*s, m, s)), lwd=2, col=col)
    lines(rep(m+3*s,2), c(0, dnorm(m+3*s, m, s)), lwd=2, col=col)
    legend("topleft", legend=paste("Std. Error = ", round(s,3)), cex=0.7, bty="n")
  }
# All you really need is a histogram in storage
  h0 <- hist(storage_b0, 
             col="skyblue3", 
             main="Sampling Distribution\n Y-intercept",
             xlab=expression(paste("Estimates of ", beta[0], " from each Sample")),
             freq=FALSE, yaxt='n', ylab="")
  m0 <- mean(storage_b0)
  s0 <- sd(storage_b0)
  addnorm(m0,s0, col="green")
  
  h1 <- hist(storage_b1, 
             col="skyblue3", 
             main="Sampling Distribution\n Slope",
             xlab=expression(paste("Estimates of ", beta[1], " from each Sample")),
             freq=FALSE, yaxt='n', ylab="")
  m1 <- mean(storage_b1)
  s1 <- sd(storage_b1)
  addnorm(m1,s1, col="green")

# y-intercept distribution tells us the highest and lowest and most imp where we usually land on yaxis

cars.lm <- lm(dist ~ speed, data=cars)
summary(cars.lm)
confint(mylm)



library(pander)
cars.lm <- lm(dist ~ speed, data=cars)
pander(summary(cars.lm)$coefficients)
```
##### Class Code t-test pvalue

[t-distrubution applet](https://byuimath.com/apps/normprobwitht.html)
The area to the left is the percentile.
3 dots gives you 1 degree of freedom. 
Linear regression for  8 dots is 6 degrees of freedom.
If we double the left value, it won't be more than half.
pt stands for percentile of t (positive gives tv over 50%)
t-value measures the size of the difference relative to the variation in your sample-- it is the difference represented in units of standarad error. A t-test puts the t-value in context of the calculated t-distribution so you can derive a probability.
tvalue how far your slop/int is from hyp one (your data)
tvalue is a calculated value derived from sample data used to test a hypothesis. It measures evidence. <br>
tstar is a critical value from the t-disribution table used to determine confidence intervals and thresholds for rejecting the null. It sets the boundardy for significance.<br>
tstar the number of standard dev from mean to capture 95.
p-value- this is what I've seen. What's the chance of that if my belief is true?
confidence interval says the truth lives there (in interval). We'll capture the truth 95% of the time.
Without standard error you have neither p-value nor conf int
Standard error for statistics (we can't say it's deviating because it's a number we don't have), standard deviation for data (we know how much our numbers deviate from each other. We don't know how much the truth deviates.)
logic of p-value is it helps you to know when you should re-evaluate your belief. You can't have probability until you have belief. If the coin was fair it has .5 chance

```{r}
# compute t
# using items from lm summary
calculate_t_value <- function(estimate, std_error, number=0){
  # number is the value from hypotheses eg. H0:B0=3)
  (estimate - number)/std_error
}

calculate_p_value <- function(t_value, dof){
  2*pt(-abs(t_value), dof)
}
calculate_p_value(19.55,397)
calculate_p_value(-6.9795,99)
-7.453-(-0.947)/2

# compute P-value use percentile function for t-dist pt(tvalue, DofF)
# gives one-tail, so multiply by 2 for alt hyp with not equal

# find p-value 2 tailed where b1=3.932, sb1=.4155, H0=3
t_value <- calculate_t_value(3.932, .4155, 3)
2*pt(-abs(t_value), 48) #p-value

# find the two-tailed p-value for cars dist~speed
t_value <- calculate_t_value(-17.58, 6.758)
t_value # -2.601
2*pt(-abs(t_value), 48) # p-value 0.0123

# for confidence intervals use confint(mylm)
# calc conf intvl for 1 or more param in fitted model. 
predict(..., interval="prediction") #predict ind value
predict(..., interval="confidence") # estimate location of mean y

curve(dt(x, 3), from=-4, to=4, lwd=2)
   curve(dnorm(x), add=TRUE, col="gray")
   abline(h=0, v=c(-1,1), col=c("gray","orange","orange"), lwd=c(1,2,2))

# What is the two-sided p-value for a t-value of 2.991 when there are 48 degrees of freedom? Or what is the two-sided probability.
pt(-abs(2.991), 48)*2 #gives the area more extreme than t=-1 for t-dist with 3 df. *2 for both sides

qt()  # quantile is a value for t that gives you a certain percentile. qt() takes percentile and gives you the tvalue. qt() & pt() are inverse. pt() takes t-value, gives percentile(area)
#Degrees of freedom 48. Change the area of the percentile to be 0.975 (the 97.5 percentile). You should get the t-value of 2.011, which can be obtained in R by the code:
qt(1-0.05/2, 48)

# Now unshade the left-tail, so just middle shaded. Area=.95. The area/percentile is the confidence value and the t-value is the critical value. Shading only the left tail is the percentile.

#What is the critical value for 90% confidence in a t distribution with 13 degrees of freedom? Show the answer using both the online applet and using R. 
qt(1-.1/2, 13)


pt(-abs(-2.601), 48)*2

#b0+-t*sd (use tstar not t-value)
#t-value relates to your data
#tstar is a choice about how confident
cars.lm <- lm(dist ~ speed, data=cars)
summary(cars.lm)
confint(cars.lm)
# use std. error of slope/int and critical value of t-disribution (tstar) to get confidence levels
# # estimate +or- tstar*stderror
3.932409 + qt(c(0.025, 0.975), 48)*0.4155 #speed

calculate_conf <- function(estimate, min=0.025, max=0.975,
                           dof, std_error){
  # min and max for qt function)
  estimate + qt(c(min, max), dof)*std_error
}
calculate_conf(3.932409, dof=48, std_error=0.4155)
confint(cars.lm)
```



##### Quizzes Hyp Param

```{r, eval=FALSE}
# Skills
# Problem 1) Install the Ecdat library in R: install.packages("Ecdat").

#From library(Ecdat) open the Caschool data set in R. As stated in the help file for this data set, this data is a collection of measurements on 420 different school districts from California during the 1998-1999 school year.

#The school districts in California offer a reduced-price lunch program. This is in a way, a measure of the poverty of the student body of the school district. We will assume that the higher the percentage of participants, the greater the general level of poverty. The question is, does the poverty level (or at least the percentage of participation in the reduced-lunch program) predict how well the student body will perform overall on a standardized test?
library(Ecdat)
?Caschool
# Equation is Y_i = B0 + B1(X_i) + epsilon_i
cas.lm <- lm(testscr~mealpct, data=Caschool)

ggplot(data=Caschool, mapping=aes(y=testscr, x=mealpct)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()

summary(cas.lm) #B0 681.43952, B1 -0.61029

#Part (c)
#Report the test statistics and p-values for the following hypotheses. (I was thinking the test statistic would be calculated, but it was the t-value)
#
#The test statistics in this case provide the number of standard errors that the estimated value of the parameter sits from the hypothesized value of the true parameter. 
   
# d, State the slope, y-intercept, and  R2 of this regression. Further, provide 95% confidence intervals for the slope and intercept. Interpret the values.
confint(cas.lm)
# b0 679.6912170 683.1878250
# b1  -0.6437314  -0.5768403
 # R2 0.7548
plot(cas.lm, which=1:2)

#problem 2
#Open the Clothing data set from library(Ecdat).

#Although this data is from 1990, it contains two interesting variables (1) the total tsales of the clothing stores and (2) the average number of hours worked per employee during the year, hourspw.

?Clothing
View(Clothing)
cl.lm <- lm(tsales~hourspw, data=Clothing)
summary(cl.lm)
ggplot(data=Clothing, mapping=aes(y=tsales, x=hourspw)) + geom_point(pch=16) +
geom_smooth(method="lm", formula=y~x, se=F) +
  theme_minimal()
#test statistic and p-value of b0=0=b1
#b0 0.026    0.979
#b1 13.218   <2e-16
#
#Now, use your own calculations to obtain test statistics and p-values for the following hypotheses.H0:B0=1500, B1=35000

#You may find useful information on how to do this in the “Explanation” tab under “t Tests” from your Math 325 Notebook, Simple Linear Regression page.
t_value <- calculate_t_value(1745, 67479, 1500)#intercept
t_value
calculate_p_value(t_value, 398)
2*pt(-abs(t_value), 48)
t_value <- calculate_t_value(43885, 3320, 35000)#slope
t_value
calculate_p_value(t_value, 398)
confint(cl.lm)
#b0 -130915.29 134404.41
#b1 37357.77  50411.97
plot(cl.lm, which=1:2)

Clothing2 <-Clothing[-397,]
cl2.lm <- lm(tsales~hourspw, data=Clothing2)
summary(cl2.lm)

#Create a residuals vs fitted-values plot and Q-Q Plot of the residuals for this regression. What do these plots show?

#They show some possible difficulties with linearity and constant variance, but more importantly observation number 397 is an extreme outlier that should be removed. This outlier is causing the estimate of the, called the MSE, to be much larger than it should be, thus causing the  R2 value to be lower than it actually should be.
# From google: The Residual Standard Error (RSE) measures the average distance that observed values fall from the regression line, expressed in the same units as the response variable. The Mean Squared Error (MSE) is the average of the squared differences between observed and predicted values, typically used for optimization as it penalizes large errors.RSE is the sqrt(MSE). RSE is in the same units as the dependent variable (Y), while MSE is in squared units (Y^2).  RSE represents the "standard deviation" of the residuals (typical prediction error). MSE is primarily a variance-based loss function to minimize. Lower fit is better for RSE, Lower fit is better for MSE.
# library(car)
boxCox(cl2.lm)
# to figure out which x transformation use the images in the text under x transformations (this needed log(x))
cl2.lm.trans <- lm(sqrt(sqrt(tsales))~log(hourspw), data=Clothing2)
b <- coef(cl2.lm.trans)
# # Plotting with transformations in aesthetics
ggplot(data=Clothing2, mapping=aes(y=sqrt(sqrt(tsales)), x=log(hourspw))) + 
  geom_point(pch=16) +
  geom_smooth(method="lm", formula=y~x, se=F) +
  #scale_y_continuous(n.breaks = 7) +
  theme_minimal()

# back in normal #don't know whether this is correct because it wasn't watching for this
ggplot(data=Clothing2, mapping=aes(y=tsales, x=hourspw)) + 
  geom_point(pch=16) +
  geom_smooth(method="lm", formula=y~x, se=F) +
  stat_function(fun = function(x) (b[1] + b[2] * log(x))^4, color = "red") +
  theme_minimal()


#supposedly this works too
ggplot(your_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", formula = I(y^0.25) ~ log(x))
ggplot(Clothing2, aes(y = tsales, x = hourspw)) +
  geom_point() +
  geom_smooth(method = "lm", formula = I(y^0.25) ~ log(x))
```

### 05 Confidence vs Prediction Intervals
Average is only only useful when spread is small.
##### Class Code Conf v Predict Int

```{r}


grid()
?faithful
predict(faith.lm, data.frame(eruptions=2))
summary(faith.lm)
# he used the rse to create a prediction interval. RSE=5.9 Two standard errors would be +- 12 minutes. Making it nice for humans +- 10 minutes which is about a 90 % confidence.

# both below have 95% confidence level
# predict tells you where you'll find the dot, if you're having a single experience you care about the prediction (buyer)
# prediction is conf + about 2 residual standard errors (our data)
# this code helps us accurately predict the next eruption
predict(faith.lm, data.frame(eruptions=2),
        interval="prediction") #wider interval, your predictions will be right more often

# confidence tells you where the true law lies
# when you'll have many experiences you care about the confidence (dealer). They care about the sum total.
predict(faith.lm, data.frame(eruptions=2),
        interval="confidence")


# se=T is giving the confidence interval
library(tidyverse)
ggplot(faithful, aes(x=eruptions, y=waiting)) + 
  geom_point( ) + 
  geom_smooth(method="lm", se=T, formula=y~x) + 
  geom_segment(x=2, xend=2, y=43.23, yend=66.63, color="orange", alpha=0.01, lwd=3)



library(mosaic)
ggplot(KidsFeet, aes(x=width, length)) + 
  geom_point() + 
  geom_smooth(method="lm", se=T, formula=y~x)
```

##### Class Code Plotting Prediction Int
```{r}
library(mosaic)
library(tidyverse)

u.lm <- lm(gasbill ~ temp, data=Utilities)
summary(u.lm)

# if you store the numbers you can index them
# [1]=fit, [2]=lwr, [3]=upr in the plot
pred <- predict(u.lm, data.frame(temp=30), 
        interval="prediction")

ggplot(Utilities, aes(x=temp, y=gasbill)) + 
  geom_point() + 
  geom_smooth(method="lm", se=T, formula=y~x) + 
  geom_segment(x=30, xend=30, y=pred[2], yend=pred[3],
               color="red", alpha=0.01, lwd=4) + 
  theme_classic()

# baseR version
# plot(gasbill ~ temp, data=Utilities)
# abline(u.lm)
# lines(c(30,30), c(83.87, 189.46), lwd=4, 
#       col=rgb(1,0,0, .5))


library(car)
boxCox(u.lm)

u.lm.t <- lm(sqrt(sqrt(gasbill)) ~ temp, data=Utilities)
b <- coef(u.lm.t)

myx <- 70
mypreds <- predict(u.lm.t, data.frame(temp=myx), 
        interval="prediction")^4


ggplot(Utilities, aes(x=temp, y=gasbill)) + 
  geom_point() + 
  geom_smooth(method="lm", se=T, formula=y~x) + 
  stat_function(fun=function(x) (b[1] + b[2]*x)^4) + 
  geom_segment(x=myx, xend=myx, y=mypreds[2], yend=mypreds[3],
               color="red", alpha=0.01, lwd=4) + 
  theme_classic()

# baseR version
# plot(gasbill ~ temp, data=Utilities)
# curve((b[1] + b[2]*x)^4, add=TRUE)
# lines(c(myx,myx), c(mypreds[2], mypreds[3]), lwd=4, 
#       col=rgb(1,0,0, .5))

```
##### Quizzes Conf Pred

```{r, eval=FALSE, fig.height=4, fig.width=4}
library(alr4)
?BGSall
View(BGSall)
htlm <- lm(HT18~HT2, data = BGSall)
summary(htlm)
plot(data=BGSall, HT18~HT2)
abline(htlm)


t_value <- calculate_t_value(45.7966, 16.700, 0)
2*pt(-abs(t_value), 134)
t_value

t_value <- calculate_t_value(1.4441, 0.1901, 2) # est, std.error, hyp value
t_value
calculate_p_value(t_value, 134) #t-value, dof

confint(htlm)
plot(which:2)

par(mfrow=c(1,3))
plot(htlm, which = 1:2) #1,3 & 2
plot(htlm$residuals) #5
boxCox(lm) # lambda for Y

predict(htlm, data.frame(HT2=83.8), interval = "prediction")

# Problem 2
# Open the wblake data set from library(alr3).
?wblake
View(wblake)

# HARD TO TELL WHICH IS X and WHICH IS Y
#If you love fishing, then you might like this data. The goal of this data was to see if there was a link in the radius size of a key “Scale” of a fish and the overall “Length” of the fish.
## this was the correct answer
# $ \underbrace{\hat{Y}_i}_\text{scale} = b_0 + b_1 \underbrace{X_i}_\text{Length}
wb.lm <- lm(Scale~Length, data=wblake)
summary(wb.lm)
ggplot(data=wblake, aes(y=Scale, x=Length)) +
  geom_point(col=rgb(.2,.2,.2,.2)) + 
  scale_y_continuous(n.breaks = 7) +
  geom_smooth(method="lm", se=T, formula=y~x)
plot(wb.lm, which=1:2)


boxCox(wb.lm)
wb.lm.sqrt <- lm(sqrt(Scale)~Length, data=wblake)
b <- coef(wb.lm.sqrt)
ggplot(data=wblake, aes(y=sqrt(Scale), x=Length)) +
  geom_point(col=rgb(.2,.2,.2,.2)) + 
  scale_y_continuous(n.breaks = 6) +
  geom_smooth(method="lm", se=T, formula=y~x)
summary(wb.lm.sqrt)
plot(wb.lm.sqrt, which=1:2)
# The answer was the diagnostic plots looked about he same, but they looked worse to me. He said he doesn't go by the red line, he squints and tries to see what the bulk of the dots are doing.
# original scale
ggplot(data=wblake, aes(y=Scale, x=Length)) +
  geom_point(col=rgb(.2,.2,.2,.2)) + 
  scale_y_continuous(n.breaks = 7) +
  scale_x_continuous(n.breaks = 4) +
  geom_smooth(method="lm", se=F, formula=y~x) +
  stat_function(fun=function(x) (b[1] + b[2]*x)^2, aes()) +
  labs(title="Radius of Key Scale Increase by Length of Fish") +
  theme_minimal()
# Where the x-axis ranges from    to   and the y-axis ranges from    to  
#Place two prediction intervals for the Scale radius when the Length of the fish is 250 on your scatterplot of the data in the original units.

#    Show the prediction interval from the original regression in the original units.
#    Show the prediction interval from the transformed regression back on the original units.
pred.o <- predict(wb.lm, data.frame(Length=250),
        interval="prediction")
pred.t <- (predict(wb.lm.sqrt, data.frame(Length=250),
        interval="prediction"))^2 # when you need the transformed prediction interval in the original units (for a plot for example), you have to take the output and do the same as you do for the plot (i.e. reverse it--Do Not do it to predict's data.frame)

ggplot(data=wblake, aes(y=Scale, x=Length)) +
  geom_point(col=rgb(.2,.2,.2,.2)) + 
  geom_smooth(method="lm", se=F, formula=y~x, color="gray") +
  stat_function(fun=function(x) (b[1] + b[2]*x)^2, color="skyblue") +
  geom_vline(xintercept = 250, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = pred.o[1], linetype = "dashed", color = "gray") +
  geom_hline(yintercept = pred.o[2], linetype = "dashed", color = "gray") +
  geom_hline(yintercept = pred.o[3], linetype = "dashed", color = "gray") +
  geom_hline(yintercept = pred.t[1], linetype = "dashed", color = "skyblue") +
  geom_hline(yintercept = pred.t[2], linetype = "dashed", color = "skyblue") +
  geom_hline(yintercept = pred.t[3], linetype = "dashed", color = "skyblue") +
  labs(title="Radius of Key Scale Increase by Length of Fish") +
  theme_minimal()


```



### 06 Different Types of Regression Models

##### Class Code: Model Types

```{r, eval=FALSE}
library(car)

lm1 <- lm(Temp ~ Month, data=airquality)

boxCox(lm1)

lm2 <- lm(Temp^2 ~ Month, data=airquality)

plot(Temp ~ Month, data=airquality)
b <- coef(lm2)
curve(sqrt(b[1] + b[2]*x), add=TRUE)

lm3 <- lm(Temp ~ Month + I(Month^2), data=airquality)
b <- coef(lm3)

curve(b[1] + b[2]*x + b[3]*x^2, add=TRUE, col='hotpink')
```

##### Quizzes: Model Types

See Skills quiz different types of models in this folder
```{R}
1
What is the equation of the regression function shown in this plot?
  ## Code to make this graphic...

> ggplot(Wages1, aes(x=school, y=wage)) + geom_point(color=rgb(.3,.3,.2,.4)) + geom_smooth(method="lm", formula= y ~ poly(x, 2, raw=TRUE), se=F, col=rgb(.3,.3,.2,.8)) + theme_bw() + labs(title="Best wages for highest schooling... or no schooling?", subtitle="Wages1 data, library(Ecdat)", x="Number of years of Schooling (school)", y="Current Hourly Wage in 1980's USD (wage)")

 

## Code needed to obtain the solution...

> summary(lm(wage ~ school + I(school^2), data=Wages1))

Call:
lm(formula = wage ~ school + I(school^2), data = Wages1)

Residuals:
Min 1Q Median 3Q Max
-7.337 -1.998 -0.474 1.444 34.554

Coefficients:
                      Estimate    Std. Error     t value    Pr(>|t|)
(Intercept)       6.13334      1.46274       4.193    2.82e-05 ***
school            -0.68354      0.25743      -2.655    0.00796 **
I(school^2)      0.05488       0.01129       4.859   1.23e-06 ***
---
Signif. codes: 0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 3.126 on 3291 degrees of freedom
Multiple R-squared: 0.08636, Adjusted R-squared: 0.0858
F-statistic: 155.5 on 2 and 3291 DF, p-value: < 2.2e-16

 

 

## Code to draw the plot in base graphics...

> plot(wage ~ school, data=Wages1, pch=16, col=rgb(.3,.3,.2,.4), main="Best wages for highest schooling... or no schooling?",  xlab="Number of years of Schooling (school)", ylab="Current Hourly Wage in 1980's USD (wage)")

> mtext(side=3, text="Wages1 data, library(Ecdat)")

> abline(v=seq(2,16,2), h=seq(0,40,10), lty=2, col=rgb(.2,.2,.2,.2))

> wages.lm <- lm(wage ~ school + I(school^2), data=Wages1)

> b <- coef(wages.lm)

> curve(b[1] + b[2]*x + b[3]*x^2, add=TRUE, col=rgb(.3,.3,.2,.8), lwd=2)

 3
 Open the Loblolly data set in R.

Which regression model shows the best linearity according to the residuals vs fitted values plot of that regression?
 plot(lm(height ~ age + I(age^2), data=Loblolly), which=1)

plot(lm(height ~ age + I(age^2) + I(age^3), data=Loblolly), which=1)

plot(lm(height ~ age + I(age^2) + I(age^3) + I(age^4), data=Loblolly), which=1)

plot(lm(height ~ age + I(age^2) + I(age^3) + I(age^4) + I(age^5), data=Loblolly), which=1)

will show the last one has the "flattest" red line, witnessing the greatest "linear" fit of the data. If you wanted to resolve the lack of constant variance, you could then apply: 

boxCox(lm(height ~ age + I(age^
                             2) + I(age^3) + I(age^4) + I(age^5), data=Loblolly))

and the resulting transformation looks quite nice:

plot(lm(sqrt(sqrt(height)) ~ age + I(age^2) + I(age^3) + I(age^4) + I(age^5), data=Loblolly), which=1)

except linearity is still lacking...

plot(lm(sqrt(sqrt(height)) ~ age + I(age^2) + I(age^3) + I(age^4) + I(age^5), data=Loblolly), which=2)

 
```

### 07 Lowess

locally weighted scatterplot smoothing--used to for quick scatterplot smoothing
The residuals v fitted plot uses a lowess line
Local Fitting: It divides data into smaller "neighborhoods." For each point, it fits a simple linear or quadratic model using only nearby data points.
Weighting: Points closer to the target point are given more "weight" in the calculation, while distant points have less influence.
Robustness: It can iteratively reduce the weight of outliers to prevent them from distorting the smooth curve.

Note that loess (modern) is for statistical modeling and prediction, uses a formula in teh interface, supports predict(), and allows up to 4 predictors

Span is the most critical param--defines the proportion of data for each local fit, lower makes it more wiggly, higher more spooth by averageing over more data, iter-the number of robustifying iterations to perform to handle outliers

In ggplot2, use geom_smooth(method = "loess") to automatically add a smooth trend line to your plots.

Lowess think of the W as a jagged line can only do degree=1, Loess can do degree=2 made with quadratics
We call this smooth line created by using biases, the unbiased regression line.

Wednesday
Has Recover the true model rmd.

##### Class Code: Lowess

```{r, eval=FALSE}
# this was from a search because I was having a hard time hearing the lecture
plot(cars$speed, cars$dist)

# Add LOWESS line (default span is 2/3)
lines(lowess(cars$speed, cars$dist), col = "red")

plot(cars$speed, cars$dist)
#### from lecture
lines(lowess(cars$speed, cars$dist, f=0.2), col = "red") # lower more wiggly
mylo <- loess(dist~speed, data=cars)
summary(mylo) #the equivalest number of params in summary gives a hint as to what might fit the data well, the span tells how much of the data in each segment

ggplot(cars, aes(x=speed, y=dist)) +
  geom_point() +
  geom_smooth()

myquad <- lm(dist ~ speed + I(speed^2) + I(speed^3), + I(speed^4), data=cars)
b <- coef(myquad)
ggplot(cars, aes(x=speed, y=dist)) +
  geom_point() +
  geom_smooth() +
  stat_function(fun=function(x) b[1] + b[2]*x + b[3]*x^2 + b[4]*x^3,
                color="green") +
  stat_function(fun=function(x) b[1] + b[2]*x + b[3]*x^2 + b[4]*x^3 + b[5]*x^4,
                color="blue")
```

```{r, eval=FALSE, fig.height=4, fig.width=4}
This was to reproduce certain graphics
X <- c(1, 4,  7,   8,  10, 20)

Y <- c(3, 5, 18, 13, 12,   1)

 

w <- c(1, .5, .2, 1, 1, .1)

mylm <- lm(Y ~ X, weights=w)

plot(Y ~ X, pch=21, bg=rgb(1-w,1-w,1-w), col="orange")

abline(mylm)
### remember curve code in base r requires a lower case x

### provided for lines
# Get some data for X and Y
X <- cars$speed
Y <- cars$dist

# Ensure the data has no missing values in any of the X,Y pairs
keep <- which(!is.na(X) & !is.na(Y))
X <- X[keep]
Y <- Y[keep]

# Decide on the fraction of points to use for the local regressions
f <- 1/2

# Identify the total sample size and sample size corresponding to f
n <- length(X) #total sample size
nn <- floor(n*f) #number of dots corresponding to percentage f

# Create storage for the lowess fitted-values
lfit <- rep(NA,n)

# Begin the for loop to compute lowess fitted-values
for (xc in 1:n){ #let each dot be the center dot, one at a time
  xdists <- X - X[xc]  #compute distance from all X-values to center dot x-value
  r <- sort(abs(xdists))[nn] #locate the nn"th" largest x-distance from center x-value
  xdists.nbrhd <- which(abs(xdists) < r) #identify x-values within the neighborhood
  w <- rep(0, length(xdists)) #initialize the weights for all x-values at 0
  w[xdists.nbrhd] <- (1 - abs(xdists[xdists.nbrhd]/r)^3)^3 #tri-cubic weight function 
  plot(Y ~ X, pch=21, bg=rgb(.53,.81,.92, w), #color dots by their weights
       col=rgb(.2,.2,.2,.3), cex=1.5, yaxt='n', xaxt='n', xlab="", ylab="")
  points(Y[xc] ~ X[xc], pch=16, col="orange") #add center dot to plot
  lmc <- lm(Y ~ X, weights=w) #run weighted regression
  curve(lmc$coef[1] + lmc$coef[2]*x, #draw line on neighborhood using from and to
        from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="orange", add=TRUE)
  lines(lfit[1:xc] ~ X[1:xc], col="gray") #add lowess line up to current center dot
  
  #lines(lowess(X,Y), col=rgb(0.698,0.133,0.133,.2))
  cat("\n\n")
  readline(prompt=paste0("Center point is point #", xc, "... Press [enter] to continue..."))
  
  
  MADnotThereYet <- TRUE
  count <- 0
  while(MADnotThereYet){ #while loop will continue until "MadnotThereYet" becomes false
    
    readline(prompt=paste0("\n   Adjusting line to account for outliers in the y-direction... Press [enter] to continue..."))   
    
    # overwrite current regression to lighter color if new regression still needed:
    curve(lmc$coef[1] + lmc$coef[2]*x, from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="wheat", add=TRUE)
    
    # update the weights
    MAD <- median(abs(lmc$res))
    resm <- lmc$res/(6*MAD)
    resm[resm>1] <- 1
    bisq <- (1-resm^2)^2
    w <- w*bisq
    obs <- coef(lmc)
    lmc <- lm(Y ~ X, weights=w)
    
    curve(lmc$coef[1] + lmc$coef[2]*x, from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="orange", add=TRUE)
    
    #stopping criterion for the weighted regressions in y-direction
    count <- count + 1
    if ( (sum(abs(obs-lmc$coef))<.1) | (count > 3))
      MADnotThereYet <- FALSE
    
  }
  
  curve(lmc$coef[1] + lmc$coef[2]*x, from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="green", add=TRUE)
  points(lmc$coef[1] + lmc$coef[2]*X[xc] ~ X[xc], pch=16, col="green")
  
  
  readline(prompt=paste0("\n   Use final line to get fitted value for this point... Press [enter] to continue to next point..."))
  
  lfit[xc] <- predict(lmc, data.frame(X=X[xc]))
  lines(lfit[1:xc] ~ X[1:xc], col="gray")
  
  
  if (xc == n){
    readline(prompt=paste0("\n  Press [enter] to see actual Lowess curve..."))
    lines(lowess(X,Y, f=f), col="firebrick")
    legend("topleft", bty="n", legend="Actual lowess Curve using lowess(...)", col="firebrick", lty=1)
  }
  
  
}


### using quadratics instead of lines
# Get some data for X and Y
X <- cars$speed
Y <- cars$dist

# Ensure the data has no missing values in any of the X,Y pairs
keep <- which(!is.na(X) & !is.na(Y))
X <- X[keep]
Y <- Y[keep]

# Decide on the fraction of points to use for the local regressions
f <- 0.5

# Identify the total sample size and sample size corresponding to f
n <- length(X) #total sample size
nn <- floor(n*f) #number of dots corresponding to percentage f

# Create storage for the lowess fitted-values
lfit <- rep(NA,n)

# Begin the for loop to compute lowess fitted-values
for (xc in 1:n){ #let each dot be the center dot, one at a time
  xdists <- X - X[xc]  #compute distance from all X-values to center dot x-value
  r <- sort(abs(xdists))[nn] #locate the nn"th" largest x-distance from center x-value
  xdists.nbrhd <- which(abs(xdists) < r) #identify x-values within the neighborhood
  w <- rep(0, length(xdists)) #initialize the weights for all x-values at 0
  w[xdists.nbrhd] <- (1 - abs(xdists[xdists.nbrhd]/r)^3)^3 #tri-cubic weight function 
  plot(Y ~ X, pch=21, bg=rgb(.53,.81,.92, w), #color dots by their weights
       col=rgb(.2,.2,.2,.3), cex=1.5, yaxt='n', xaxt='n', xlab="", ylab="")
  points(Y[xc] ~ X[xc], pch=16, col="orange") #add center dot to plot
  lmc <- lm(Y ~ X, weights=w) #run weighted regression
  curve(lmc$coef[1] + lmc$coef[2]*x, #draw line on neighborhood using from and to
        from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="orange", add=TRUE)
  lines(lfit[1:xc] ~ X[1:xc], col="gray") #add lowess line up to current center dot
  
  #lines(lowess(X,Y), col=rgb(0.698,0.133,0.133,.2))
  cat("\n\n")
  readline(prompt=paste0("Center point is point #", xc, "... Press [enter] to continue..."))
  
  
  MADnotThereYet <- TRUE
  count <- 0
  while(MADnotThereYet){ #while loop will continue until "MadnotThereYet" becomes false
    
    readline(prompt=paste0("\n   Adjusting line to account for outliers in the y-direction... Press [enter] to continue..."))   
    
    # overwrite current regression to lighter color if new regression still needed:
    curve(lmc$coef[1] + lmc$coef[2]*x, from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="wheat", add=TRUE)
    
    # update the weights
    MAD <- median(abs(lmc$res))
    resm <- lmc$res/(6*MAD)
    resm[resm>1] <- 1
    bisq <- (1-resm^2)^2
    w <- w*bisq
    obs <- coef(lmc)
    lmc <- lm(Y ~ X, weights=w)
    
    curve(lmc$coef[1] + lmc$coef[2]*x, from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="orange", add=TRUE)
    
    #stopping criterion for the weighted regressions in y-direction
    count <- count + 1
    if ( (sum(abs(obs-lmc$coef))<.1) | (count > 3))
      MADnotThereYet <- FALSE
    
  }
  
  curve(lmc$coef[1] + lmc$coef[2]*x, from=min(X[xdists.nbrhd]), to=max(X[xdists.nbrhd]), col="green", add=TRUE)
  points(lmc$coef[1] + lmc$coef[2]*X[xc] ~ X[xc], pch=16, col="green")
  
  
  readline(prompt=paste0("\n   Use final line to get fitted value for this point... Press [enter] to continue to next point..."))
  
  lfit[xc] <- predict(lmc, data.frame(X=X[xc]))
  lines(lfit[1:xc] ~ X[1:xc], col="gray")
  
  
  if (xc == n){
    readline(prompt=paste0("\n  Press [enter] to see actual Lowess curve..."))
    lines(lowess(X,Y, f=f), col="firebrick")
    legend("topleft", bty="n", legend="Actual lowess Curve using lowess(...)", col="firebrick", lty=1)
  }
  
  
}

```
##### Quizzes

The rest of the Lowess code is in broken

```{r}


```



### 11/12 Logistic Regression

b0 controls where crosses y axis but when set to 0 example crosses at .5, the b1 is controlling left to right shift, rare to get the nice curve on real data

Our model gives probabilities of y being 1

```{r}
set.seed(1)
n <- 100
beta0 <- -10
beta1 <- 2

x <- runif(n,0,10)
prob <- exp(beta0+beta1*x)/(1+exp(beta0+beta1*x))

y <- rbinom(n,1, prob) # coin toss

plot(y~x, col=rgb(0,1,0, 0.1), pch=16, cex=2)

myglm <- glm(y~x, family=binomial)
summary(myglm)
b <- coef(myglm)
curve(exp(b[1] + b[2]*x)/(1+exp(b[1]+b[2]*x)), add=TRUE) # regression
curve(exp(beta0 + beta1*x)/(1+exp(beta0+beta1*x)), add=TRUE, lty=2) #true line

#quiz code: simplyamazing

# you could do this whole thing with a b2*x^2 will create a u shape 
# if the squared term is negative the quadratic (u) faces down, the b1 changes slope, shift -b1/2*b2 for the vertex of u
# you can do the b0 + b1 twice (b0 + b_1x + b_2x_2 + b_2xx_2) for two lines
```
##### Quizzes logistic regression
look at challenger example in 



```{r, fig.height=10, fig.width=10}
library(tidyverse)
library(mosaicData)
??KidsFeet
#Use the KidsFeet data set from library(mosaic) to practice fitting a logistic regression.
# Is birthmonth enough information to predict the birthyear of a fourth grade student? Fit a logistic regression of birthyear == 88 on the birthmonth of a child to find out.
# p(birthyear_i=88|birthmonth_i) = e^(B0+B1birthmonth)/(1+eB0+B1birthmonth)

#Enter the values of your estimates for  β0
  and  β1
 in the logistic model:
KidsFeet2 <- KidsFeet %>%
  mutate(by88 = ifelse(birthyear == 88, 1, 0))
View(KidsFeet2)
#Fit a new model using only the width of the child's foot instead of birthmonth to predict the birthyear of the child. Is foot width a better or worse predictor of the birthyear of the child than birthmonth?

#State the AIC, null deviance, and residual deviance of both logistic regression models.
glm1 <- glm(by88 ~ birthmonth, data=KidsFeet2, family=binomial)
summary(glm1)
#Use the better model of the two models from parts (a) and (b) to predict the probability that a fourth grade boy (from the same era as the students in this data) was born in 88 given they had a foot width of 9.5 cm, a foot length of 24, and was born in May. (This is more information than you need to make the prediction.)
predict(glm1, data.frame(birthmonth=5), type="response")

glm2 <- glm(by88 ~ width, data=KidsFeet2, family=binomial)
summary(glm2)
predict(glm2, data.frame(width=9.5))

glm1 <- glm(by88 ~ birthmonth, data=KidsFeet2, family=binomial)
summary(glm1)
Interpret the effect on the odds that the slope term from the model used in Part (c) has on the odds of a fourth grade child being born in 88.
 

#  Answer: "The odds of the child being born in 88 decrease by   
63.18261
 % for each one month increase in the birthmonth (1 being January and 12 being December). In other words, a child born in February has an odds of being born in 88 that is   
63.18261
% less than a child that was born in January. And the same is true for any two consecutive months."

 

exp(-0.9992) e^b1 
#a multiplier the odds

1-0.3681739 but then 1-the for that one



#2
library(mosaic)
#Consider the `?RailTrail` data set in library(mosaicData). As shown by the boxplot below, the number of users on the rail trail seems to be influenced by whether or not there was any precipitation on that day.
boxplot(volume ~ precip>0, data=RailTrail, col="skyblue", xlab="90 Days of Records from April to November", ylab="Total Number of Users on the Trail that Day", main="Rain Seems to Reduce the Number of Users", names = c("Days with No Rain", "Rainy Days"))


#The goal of this question is to identify a logistic regression model that could be used to predict if there will be rain on a given day or not.

#Run the following commands to reduce the data to variables of interest for this problem.
library(dplyr)
View(RT)

##ggplot needs 1s and 0s but base R can do FALSE TRUE
RT <- RailTrail %>% 
  mutate(precipOccurred = (precip > 0)) %>%
  select(precipOccurred, lowtemp, spring, summer, fall, cloudcover, weekday)

pairs(RT)

#Fit three logistic regression models. The first should use lowtemp to predict precipitation. The second should use cloudcover to predict precipitation. The third should use fall to predict precipitation.

Compare the AIC and residual deviance of each model, which one is "better"?

 

Note that each model has a null deviance of 113.14 because that all use the same y-variable and have no missing values in the x-variables. It is wise to check that the null deviance of models you are comparing match up, otherwise, their residual deviances and AIC values aren't directly comparable.
glm.lt <- glm(precipOccurred ~ lowtemp, data=RT, family=binomial)
summary(glm.lt)


glm.cc <- glm(precipOccurred ~ cloudcover, data=RT, family=binomial)
summary(glm.cc)

glm.f <- glm(precipOccurred ~ fall, data=RT, family=binomial)
summary(glm.f)

predict(glm1, data.frame(birthmonth=5), type="response")

Try every possible 2-variable (no interaction term) logistic regression model that uses both (1) the best variable from Part (b), and (2) each of the other variables in the RT data set, one at a time. Note the AIC of each model. The best two-variable model for this data is has an AIC of 82.75 and is given by:

glm.clt <- glm(precipOccurred ~ cloudcover + lowtemp, data=RT, family=binomial)
summary(glm.clt)

glm.csp <- glm(precipOccurred ~ cloudcover + spring, data=RT, family=binomial)
summary(glm.csp)

glm.csm <- glm(precipOccurred ~ cloudcover + summer, data=RT, family=binomial)
summary(glm.csm)

glm.clf <- glm(precipOccurred ~ cloudcover + fall, data=RT, family=binomial)
summary(glm.clf)

glm.cwd <- glm(precipOccurred ~ cloudcover + weekday, data=RT, family=binomial)
summary(glm.cwd)

However, the p-value for the   
weekday
  variable is not significant in this model. Also, it is interesting to note that the AIC of this mode is   
  than the one-variable model that just uses   
cloudcover
 to predict precipitation occurring. This is a good witness that the AIC helps pretect against over-fitting models.

 
```
null is like ssto, residual deviance is like ssr
<br>
### 14 
standard error hint "residual standard error"
1 Residuals (normality of residuals is the least of our problems) This section can tell us whether there's skew, the median should be around zero (the mean is always 0 for an lm)
2 when x is 0
3 change in the average y The regression line is the average y. We know only about the average we will miss because the dots are almost never on the line,
4 measures the sample distributions standard deviation, all samples vary. This is a guess as to how much variation all samples would have. The saving grace is the standard error, finding common ground. If you feel a deep sense of love for these numbers you're starting to understand statistics.
5
6 number of standard errors that the estimate is the null
7
8 t=b0-B0/S_b0
9
10 We expect the dots to vary by about 15pts on average from the line. This accounts for the individuality of the points. (I'm always wrong but I'm close.)
11
12 R2 It's the proportion of variance in Y explained by the data. (SSE is the unexplained variance)
13 Adjusted R-squared has a penalty 1 - (n-1) SSE/ (n-p)SSTO a 1 param model lmm(y~1) If you use a lot of parameters and get an adj R2 that's good. You're doing well because it cancelled out the penality
14 last is the model as the whole doing better than y-bar
15
16 A quadratic has I think 3 terms
17 bottom row is only helpful in multiple regression
p-value in summarries assuming all other variables are sig and meaningful. Depends on the context of what's present

remind how to find things in pair plots
go back to two theory assignments, skills quizzes, practice test, 
```{r}
Linear Regression Summary


```

## Code notes

##### Generic
Ctrl + L to clear console
cex=character extension
Getwd()<br>
Glimpse() to see the data and it's types (good for when ? Doesn't work)<br>
In rmd ctr shft c to comment/uncomment<br>
Echo = false means the code won't show just output<br>
Double question marks tells you where<br>
Don't put install in the script cause it'll do it every time<br>
Say no to saving workspace instead save a script with everything.<br>
Shortcut for pipe is ctrl + shift + m<br>

##### Graphics
Notebook add links to rmd files (can also put line number)<br>
P + theme(text=element_text(size24))

##### Problems
Mismatched directories can sometimes cause you problems when you knit.<br> 
Session>Set Working Directory to Source file location<br>
NAs are contagious they will permeate your data, pass na.rm true to remove the nas<br>
Session > restart R because he overwrote his data<br>
Clicking on broom cleans out all the data<br>

##### Reminders
Don't do anything to the original data file make sure to assign it to a new name!!!<br>
Filter rows/select columns<br>
Colnames() to see the column names<br>
Tibble() see code for names and first few rows<br>
Easier to do test questions in base R<br>
Filter into a new data set before summarizing<br>

<a id=" Code Notes"></a>

#### applr
get_inverse_function() 
This function analyzes the Left-Hand Side (LHS) of a formula and constructs
a function that performs the mathematical inverse.
takes A formula (e.g., log(y) ~ x) or an expression/call.
returns
A function accepting one argument (x) that returns the back-transformed value.

diagnose()
Generates a set of three diagnostic plots for a linear model object:
Residuals vs Fitted, Normal Q-Q, and a plot of the residuals.

lm_model <- lm(mpg ~ wt, data = mtcars)
diagnose(lm_model)

lm_equation()
Takes a fitted linear model and returns a human-readable equation string showing
the relationship between the response variable and predictors. This is useful for
presenting model results in reports or understanding the mathematical form of the model.
Coefficients are rounded to 3 significant figures for readability.
model <- lm(mpg ~ disp + hp, data = mtcars)

lm_latex()
Takes a fitted linear model and generates a properly formatted LaTeX equation
suitable for inclusion in R Markdown documents, academic papers, or presentations.
The output includes underbrace notation to clearly label predicted values and
predictor variables, making it ideal for educational or presentation purposes.
Coefficients are rounded to 3 significant figures.
model <- lm(width ~ length + I(length^2) + sex + sex:length + sex:I(length^2), KidsFeet)
lm_latex(model)

theme_lc() # based of Lauren Condie's work
A custom ggplot2 theme that provides a cohesive, aesthetically pleasing color scheme
and styling for statistical graphics. This theme sets both default geom aesthetics
and plot theme elements to create a consistent visual style across all plots.
The color palette uses warm, earthy tones that work well for academic and professional
presentations.

scatter_3d()
Creates an interactive 3D visualization of a linear model with exactly 2 predictor variables.
The function plots the original data points as scattered points in 3D space and overlays
a semi-transparent regression surface showing the model predictions across the entire
predictor space. This provides an intuitive way to understand how a linear model
behaves across different combinations of two predictor variables.

model <- lm(mpg ~ disp + hp, data = mtcars)
scatter_3d(model, colors=c('blue','yellow'))
lm_equation(model)

geom_slice()
\arguments{
\item{model}{A linear model provided by 'lm()'}

\item{n}{The number of predicted points across the graph - default 100}

\item{inherit.aes}{Inherits the aes provided by 'ggplot' object}

\item{predict_vars}{A list of specified variables and values that specify the slice - not required}

\item{back_transform}{Optional. Either NULL (auto-detect), a function for back-transforming predictions,
or a character string ("log", "log10", "log2", "sqrt", "exp", "inverse") specifying the transformation
to reverse. Use this when the model has a transformed response variable (e.g., log(y)) but the plot
displays the original scale.}

\item{...}{Other variables (color, alpha, etc.)}
}
\value{
A ggplot layer that displays the slice of the high dimensional model
}
\description{
A 'ggplot' layer that takes a high dimensional model and displays the predicted
model across each facet of a graph. This is the main user-facing function for
adding model slices to ggplot2 graphs.
}
\examples{
\dontrun{
# Example: Using geom_slice without specifying 'predict_vars'
model <- lm(mpg ~ disp + hp, data = mtcars)
ggplot(mtcars, aes(x=disp,y=mpg))+
geom_point()+
facet_wrap(~hp)+
geom_slice(model=model)

# Example: Using geom_slice with 'predict_vars'
model <- lm(mpg ~ disp + hp + gear, data = mtcars)
ggplot(mtcars, aes(x=disp, y=mpg))+
geom_point()+
facet_wrap(~hp)+
geom_slice(model=model, predict_vars=list(gear=4))

# Example: Model with log transformation (auto-detected)
model <- lm(log(mpg) ~ disp + hp, data = mtcars)
ggplot(mtcars, aes(x=disp, y=mpg))+
geom_point()+
geom_slice(model=model, predict_vars=list(hp=110))

# Example: Manual back-transformation specification
model <- lm(log(mpg) ~ disp + hp, data = mtcars)
ggplot(mtcars, aes(x=disp, y=mpg))+
geom_point()+
geom_slice(model=model, back_transform="log")
}

```{r, eval=FALSE, fig.height=4, fig.width=4}
model <- lm(mpg ~ disp + hp + am, data = mtcars)
slice_2d(model, xaxis = "disp")

library(ggplot2)
model <- lm(mpg ~ disp + hp + cyl, data = mtcars)
ggplot(mtcars, aes(disp, mpg)) +
  geom_point() +
  facet_wrap(~cyl) +
  geom_slice(model = model, predict_vars = list(hp = 110))

```

 Plot a 3D-grapheable linear model
#'
#' Creates an interactive 3D visualization of a linear model with exactly 2 predictor variables.
#' The function plots the original data points as scattered points in 3D space and overlays
#' a semi-transparent regression surface showing the model predictions across the entire
#' predictor space. This provides an intuitive way to understand how a linear model
#' behaves across different combinations of two predictor variables.
#'
#' @param model A saved linear model with 2 predictor variables
#' @param n Number of evaluations for creating the surface grid
#' @param colors Colors used for the gradient scale
#'
#' @returns A 3D graph in the viewer section
#'
#'
#' @examples
#' \dontrun{
#' model <- lm(mpg ~ disp + hp, data = mtcars)
#' scatter_3d(model, colors=c('blue','yellow'))
#' }
#'
#' @importFrom tidyr pivot_wider
#' @importFrom plotly plot_ly add_markers add_trace
#' @importFrom graphics lines
#' @importFrom stats coef predict terms setNames formula predict.lm
#'
#' @export
scatter_3d <- function(model, n=100, colors = c('blue', 'yellow')){

  # Validate that the input is a linear model
  if (!inherits(model, 'lm'))
    stop('Model must be in lm format')

  # Extract all variable names from the model formula
  # This includes the response variable and all predictors
  vars <- all.vars(formula(model))

  # Check that model has exactly 3 variables (1 response + 2 predictors)
  # More than 3 variables cannot be meaningfully displayed in 3D
  if(length(vars) > 3)
    stop('Not 3D Graphable, Reduce lm to 2 predictor variables')

  # Extract the data for response and predictor variables from the fitted model
  # The model$model data frame contains the original data used to fit the model
  y <- model$model[1][[1]]        # Response variable (e.g., mpg)
  x1 <- model$model[2][[1]]       # First predictor (e.g., disp)
  x2 <- model$model[3][[1]]       # Second predictor (e.g., hp)

  # Get variable names for labeling axes
  y_name <- names(model$model)[1]   # Response variable name
  x1_name <- names(model$model)[2]  # First predictor name
  x2_name <- names(model$model)[3]  # Second predictor name

  # Create a clean data frame with proper column names for plotting
  df <- setNames(data.frame(y,x1,x2), c(y_name, x1_name, x2_name))

  # Ensure x1 is numeric for creating the 3D surface
  # Convert factors to numeric values if necessary
  if (!is.numeric(x1)) {
    if (!is.factor(x1)) x1 <- as.factor(x1)  # Convert to factor first if character
    x1 <- as.numeric(x1)                      # Then convert to numeric
  }

  # Ensure x2 is numeric for creating the 3D surface
  # Convert factors to numeric values if necessary
  if (!is.numeric(x2)) {
    if (!is.factor(x2)) x2 <- as.factor(x2)  # Convert to factor first if character
    x2 <- as.numeric(x2)                      # Then convert to numeric
  }

  # Create sequences of values across the range of each predictor
  # These will form a grid for generating the prediction surface
  axisx <- seq(min(x1), max(x1), length.out=n)  # x1 values for surface
  axisy <- seq(min(x2), max(x2), length.out=n)  # x2 values for surface

  # Create a grid of all combinations of x1 and x2 values
  # expand.grid creates a data frame with every combination of the input vectors
  surface <- expand.grid(setNames(list(axisx, axisy), c(x1_name,x2_name)), KEEP.OUT.ATTRS = F)

  # Ensure factor levels are preserved if original variables were factors
  # This is important for proper prediction with categorical variables
  if (is.factor(x1))
    surface[[x1_name]] <- factor(surface[[x1_name]], levels = levels(x1))

  if (is.factor(x2))
    surface[[x2_name]] <- factor(surface[[x2_name]], levels = levels(x2))

  # Generate predictions for every point on the surface grid
  # This creates the Z-values (height) for the 3D surface
  surface$Z <- predict.lm(model, newdata = surface)

  # Reshape the surface data from long format to wide format
  # This creates a matrix suitable for plotly's surface plotting
  # Each row represents a y-axis value, each column an x-axis value
  widesurface <- pivot_wider(
    surface,
    names_from = !!sym(x1_name),      # Columns will be x1 values
    values_from = Z                    # Cell values will be predictions
  )

  # Convert to matrix format and set row names for the surface plot
  surface <- as.matrix(widesurface[,-1])           # Remove first column (x2 values)
  rownames(surface) <- widesurface[[x2_name]]     # Set row names as x2 values

  # Create the 3D plot using plotly
  # This uses dynamic code generation to handle variable names properly
  eval(parse(text=paste0("plot_ly() %>%",
    # Add the original data points as 3D scatter points
    "add_markers(data=df,",
         "x=~",x1_name,",",              # x-axis: first predictor
         "y=~",x2_name,",",              # y-axis: second predictor  
         "z=~",y_name,",",               # z-axis: response variable
         "type='scatter3d',",
         "mode='markers',",
         "colors = colors) %>%",
    # Add the regression surface
    "add_trace(z=surface,",             # z-values: predicted responses
              "x=axisx,",               # x-values: first predictor range
              "y=axisy,",               # y-values: second predictor range
              "type='surface',",
              "opacity = 0.6)")))       # Semi-transparent surface
}



#' 2D Slicer
#'
#' Take a linear model and display a 2D slice in base R. This function creates
#' a new plot comparing the response y variable with one predictor x variable
#' and displaying a linear model line with all other predictor variables
#' held at constant values. For unspecified variables, the xaxis is assumed
#' to be the first predictor in the model and all other variable are imputed
#' (mean for numeric, most common for factors).
#'
#' @param model A linear model
#' @param xaxis A non-required specification of the xaxis variable; default is first x var in lm
#' @param n Number of displayed points; default is 100
#' @param caption Boolean to include caption with values of predictor variables not shown in the chart, TRUE by default
#' @param back_transform Either 1) A function to back transform the predicted y values and interval bounds (e.g., `exp` or `\(x) exp(x)` for log-transformed models), or 2) a boolean stating whether it should apply a transformation or not (`FALSE` turns off back_transform). In most cases, this does not need to be specified as the function infers the transformation needed reliably.
#' @param ... Remaining vars (color, linetype, etc.)
#'
#' @returns A 2D graph of a sliced model
#'
#' @importFrom graphics lines
#' @importFrom stats coef predict terms setNames formula predict.lm
#'
#' @examples
#' \dontrun{
#' model <- lm(mpg ~ disp, data = mtcars)
#' slice_2d(model, xaxis = 'disp')
#' }
#'
#' @export
slice_2d <- function(model,xaxis=NULL,n=100,caption=TRUE, back_transform=TRUE, ...){

  # Prepare data
  prep <- prepare_slice_data(model, xaxis, n, caption, back_transform, ...)

  orig_data <- prep$orig_data
  new_data <- prep$new_data
  xaxis <- prep$xaxis
  y_name <- prep$y_name
  plot_args <- prep$plot_args
  caption_text <- prep$caption_text

  # PLOT

  # Create the base plot with original data points
  plot(orig_data[[y_name]] ~ orig_data[[xaxis]],
       xlab = xaxis,
       ylab = y_name,
       main = deparse(formula(model)),
       pch = 19,
       col = "steelblue")

  # Add the prediction line to show the model slice
  # do.call used for lines function to allow a list "plot_args" in place of ...
  do.call(lines, c(list(x = new_data[[xaxis]], y = new_data$preds), plot_args))

  # Add a caption with predictor values that aren't visible in the plot
  # While this is functional to some extent, it has a clear complex problem,
  # which is detailed at the bottom of the file.
  if (caption) {
    # Add the caption to the plot
    mtext(caption_text, # text
          side = 3, # Top margin
          line = 0.5, # Just under the title (line 0.5)
          adj = 0, # Left-align
          # cex = 0.8, # Smaller text size
          col = "gray40")
  }
}


#' 2D Slice to R Plot
#'
#' Take a slice from a linear model and add the displayed graph to a preexisting R plot.
#' This function is similar to slice_2d() but adds a line to an existing plot rather
#' than creating a new one. Useful for overlaying multiple model slices or adding
#' model predictions to scatter plots.
#'
#' @param model A linear model
#' @param xaxis A non-required specification of the xaxis variable; default is first x var in lm
#' @param n Number of displayed points; default is 100
#' @param caption Boolean to include caption with values of predictor variables not shown in the chart, FALSE by default
#' @param back_transform Either 1) A function to back transform the predicted y values and interval bounds (e.g., `exp` or `\(x) exp(x)` for log-transformed models), or 2) a boolean stating whether it should apply a transformation or not (`FALSE` turns off back_transform). In most cases, this does not need to be specified as the function infers the transformation needed reliably.
#' @param ... Remaining vars (color, linetype, etc.)
#'
#' @returns A line added to a preexisting R plot displaying a 2D graphable slice of an HD model
#'
#' @importFrom graphics lines
#' @importFrom stats coef predict terms setNames formula predict.lm
#'
#' @examples
#' \dontrun{
#' model <- lm(mpg ~ disp + hp, data = mtcars)
#' plot(mpg ~ disp, data = mtcars)
#' add_slice_2d(model,xaxis='disp', hp=100)
#' }
#'
#' @export
add_slice_2d <- function(model,xaxis=NA,n=100,caption=FALSE, back_transform=TRUE, ...){

  # Prepare data
  prep <- prepare_slice_data(model, xaxis, n, caption, back_transform, ...)

  new_data <- prep$new_data
  xaxis <- prep$xaxis
  plot_args <- prep$plot_args
  caption_text <- prep$caption_text

  # Warn if the slice is outside the current plot window
  usr <- try(par("usr"), silent = TRUE)
  if (!inherits(usr, "try-error") && is.numeric(usr)) {
    xlim <- usr[1:2]; ylim <- usr[3:4]
    xr <- range(new_data[[xaxis]], na.rm = TRUE)
    yr <- range(new_data$preds, na.rm = TRUE)
    if (xr[2] < xlim[1] || xr[1] > xlim[2] || yr[2] < ylim[1] || yr[1] > ylim[2]) {
      warning("Slice is outside the current plot range; the line may not be visible or aligned.")
    }
  }

  # Add the prediction line to the existing plot
  # Note: This assumes a plot already exists - it will error if no plot is active
  # do.call used for lines function to allow a list "plot_args" in place of ...
  do.call(lines, c(list(x = new_data[[xaxis]], y = new_data$preds), plot_args))

  # Add a caption with predictor values that aren't visible in the plot
  if (caption) {
    # Add the caption to the plot
    mtext(caption_text, # text
          side = 3, # Top margin
          line = 0.5, # Just under the title (line 0.5)
          adj = 0, # Left-align
          # cex = 0.8, # Smaller text size
          col = "gray40")
  }
}


# This data preparation is shared by both slice_2d and add_slice_2d
prepare_slice_data <- function(model, xaxis, n, caption, back_transform=TRUE, ...) {
  # Validate that the input is a linear model
  if (!inherits(model, 'lm')) {
    stop("Model must be in lm() format")
  }

  # Check for '$' in variable names
  for (name in names(model$model)) {
    if (grepl("\\$", name)) stop("Cannot use lm with '$' in its variable names")
  }

  # Get original variable names
  var_names <- all.vars(formula(model))
  y_name <- var_names[1]

  # Determine xaxis if not specified
  if (is.null(xaxis) || is.na(xaxis)) {
    xaxis <- var_names[2]  # Position 1 is the response variable
    message(paste0("X-axis not specified - Used first x variable as x-axis",
                   "\n    To specify, use 'xaxis = \"var_name\"' such as 'xaxis = \"",
                   xaxis, "\"'"))
  }

  # Validate that the specified x-axis variable exists in the model
  if (!(xaxis %in% var_names)) {
    stop(paste0("xaxis variable '", xaxis, "' not found in the model."))
  }

  # Extract original data
  if (!is.null(model$call$data)) {
    orig_data <- eval(model$call$data, envir = environment(formula(model)))
  } else {
    orig_data <- data.frame(
      sapply(var_names, FUN = \(v) get(v, environment(formula(model))))
    )
  }

  # Validate x-axis variable is numeric
  if (!is.numeric(orig_data[[xaxis]])) {
    stop(paste0("`xaxis` must be numeric; the class \"", class(orig_data[[xaxis]]), "\" is not supported."))
  }

  # Find all other variables in the model
  other_vars <- setdiff(var_names, c(y_name, xaxis))

  # Get any arguments passed via ...
  dots <- list(...)

  # For each "other" variable, determine what value to hold it at
  other_vals <- lapply(other_vars, function(var) {
    if (!is.null(dots[[var]])) {
      dots[[var]]
    } else {
      col <- orig_data[[var]]
      if (is.numeric(col)) {
        m <- mean(col, na.rm = TRUE)
        message(paste0("Value for `", var, "` not specified -",
                       " Used mean: ", round(m, 1),
                       "\n    To specify, use 'var_name = value', such as '",
                       var, " = ", round(m, 1), "'"))
        m
      } else if (is.factor(col)) {
        l <- levels(col)[1]
        message(paste0("Value for `", var, "` not specified -",
                       " Used first value: ", l,
                       "\n    To specify, use 'var_name = value', such as '",
                       var, " = ", l, "'"))
        l
      } else {
        stop(paste0("Unsupported variable type for: `", var, "`"))
      }
    }
  })

  # Convert the constant values into a data frame format
  x_names <- as.data.frame(as.list(setNames(other_vals, other_vars)))

  # Create an evenly spaced sequence of values across the range of the x axis
  x_vals <- seq(min(orig_data[[xaxis]], na.rm = TRUE), max(orig_data[[xaxis]], na.rm = TRUE), length.out = n)

  # Prepare x values for the predict() function
  new_data <- setNames(data.frame(x_vals), xaxis)

  # Add the constant values for all other variables
  if (length(other_vars) > 0) {
    new_data <- cbind(new_data, x_names)
  }

  # Create y-values for all the new data points
  new_data$preds <- predict(model, newdata = new_data)

  # Get y transformation
  # If they say no, don't transform
  if (is.logical(back_transform) && back_transform == FALSE) {
    inverse <- identity

  # If they specify a valid function, go with that
  } else if (is.function(back_transform) && length(formals(back_transform)) == 1) {
    inverse <- back_transform

  # Otherwise, infer the back transformation
  } else if (is.logical(back_transform) && back_transform == TRUE) {
    inverse <- get_inverse_function(formula(model))

  # And if they specified something invalid, warn them
  } else {
    warning(paste0("The back_transform argument`", back_transform,
                  "` is not a valid function or boolean.",
                  "\n    Inferring transformation from lm."))
    inverse <- get_inverse_function(formula(model))
  }

  # Apply transformation
  new_data$preds <- inverse(new_data$preds)

  # Separate model values from plotting arguments in ...
  plot_args <- dots[!names(dots) %in% other_vars]

  # Generate caption
  caption_text <- ""
  if (caption) {
    for (i in seq_along(other_vars)) {
      if (i > 1) {
        caption_text <- paste0(caption_text, "; ")
      }
      if (is.numeric(other_vals[[i]])) {
        caption_text <- paste0(caption_text, other_vars[i], ": ", round(other_vals[[i]], 1))
      } else {
        caption_text <- paste0(caption_text, other_vars[i], ": ", other_vals[i])
      }
    }
  }

  list(
    orig_data = orig_data,
    new_data = new_data,
    xaxis = xaxis,
    y_name = y_name,
    plot_args = plot_args,
    caption_text = caption_text,
    other_vars = other_vars,
    other_vals = other_vals
  )
}



# On captioning
  # Because Base R uses a "painter" system, subsequent text appears on top of
  # the previous text caption. One solution is place future captions below the
  # previous, but the plot size can't be adjusted (it was already painted).
  # The next idea is to access the previous caption and update it. However,
  # it is simply drawn on the plot and not accessible as an object. So, it
  # could be made into an object and each caption an environment variable. This
  # has two more problems. It doesn't reset with a new plot, and therefore will
  # bleed into future plots. Additionally, a white rectangle needs to be
  # painted on top of the existing text to accommodate the new text.
  # Possible probably. Complicated definitely.
  #
  # By default it is on for slice_2d, but off for add_slice_2d, because slice_2d
  # creates a fresh plot with nothing to overwrite.