---
title: "06_DifferentModels"
output: html_document
date: "2026-02-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(car)
```

## Model Selection

```{r}
lm1 <- lm(Temp ~ Month, data=airquality) 

# transform
boxCox(lm1) # says square or nothing
lm2 <- lm(Temp^2 ~ Month, data=airquality)

plot(Temp ~ Month, data=airquality)
b <- coef(lm2)
curve(sqrt(b[1] + b[2]*x), add=TRUE, col="skyblue")

# model selection
lm3 <- lm(Temp ~ Month + I(Month^2), data=airquality)
b <- coef(lm3)

curve(b[1] + b[2]*x + b[3]*x^2, add=TRUE, col="hotpink")
```

## How the quadratic model works

[Desmos](https://www.desmos.com/calculator)
Beta1 controls the vertex by -Beta1/(2*Beta2)
Beta2 controls concavity and steepness
y=1 constant, y=-2x slope, y=x^2 quadratic
Set the b0 first.
b1 rocks the parabolas location in a parabolic form
b2 is the only one that can open the parabola wider, negative flips it
For cubic, b1 has a symmetric effect (slope of middle line portion), b2 kinda width but asymetric, b3 some spread and flip
For the two lines, the x2 has to sample from 0s and 1s

```{r, echo=FALSE}
# simulating the quadratic model
#2.4, 0.8, -.09 (-6,10)
#  ## Simulating Data from a Regression Model
  ## This R-chunk is meant to be played in your R Console.
  ## It allows you to explore how the various elements
  ## of the regression model combine together to "create"
  ## data and then use the data to "re-create" the line.

  set.seed(101) #Allows us to always get the same "random" sample
                #Change to a new number to get a new sample

  n <- 100 #set the sample size

  X_i <- runif(n, -6, 10) 
    #Gives n random values from a uniform distribution between 15 to 45.

  beta0 <- 2.4 #Our choice for the y-intercept. 

  beta1 <- 0.8 #Our choice for the slope. 
  
  beta2 <- -0.09

  sigma <- .46 #Our choice for the std. deviation of the error terms.


  epsilon_i <- rnorm(n, 0, sigma) 
    #Gives n random values from a normal distribution with mean = 0, st. dev. = sigma.

  Y_i <- beta0 + beta1*X_i + beta2*X_i^2 + epsilon_i 
    #Create Y using the normal error regression model

  fabData <- data.frame(y=Y_i, x=X_i) 
    #Store the data as data

  View(fabData) 
  

  #In the real world, we begin with data (like fabData) and try to recover the model that 
  # (we assume) was used to created it.

  fab.lm <- lm(y ~ x + I(x^2), data=fabData) #Fit an estimated regression model to the fabData.
  b <- coef(fab.lm)
  summary(fab.lm) #Summarize your model. 

  plot(y ~ x, data=fabData, pch=16) #Plot the data.

  #abline(fab.lm) #Add the estimated regression line to your plot.


  # Now for something you can't do in real life... but since we created the data...

  #abline(beta0, beta1, lty=2) 
  curve(beta0 + beta1*x + beta2*x^2, add=TRUE, lty=2)
  curve(b[1]+b[2]*x+b[3]*x^2, add=TRUE)
    #Add the true regression line to your plot using a dashed line (lty=2). 

  legend("topleft", legend=c("True Line", "Estimated Line"), lty=c(2,1), bty="n") 
    #Add a legend to your plot specifying which line is which.
```

The quadratic model should be concave down.
The maximum point (the vertex) should occur in the upper-right hand side of the plot.
The quadratic model should exit the graph in both the bottom-left corner and the middle of the right-hand side.
The number of dots in your plot should be similar to the number of dots in this plot, but don't worry about counting exactly.
The spread of the dots in your plot should be similar to the spread of the dots in this plot.
Your plot should have a legend of some sort, though you don't have to use the fancy math symbols that this one did. (Hint: Try ?plotmath if you want to try the symbols though.)

```{r}
challData <- data.frame(y=c( 62.0230141814838,65.3287607440202,63.0850996714816,64.7215795301002,63.5747270663006,53.702998209753,67.3864948748882,63.6127706891815,67.7168411573423,65.1905587130745,62.7203672543605,65.9786946907104,65.9747299777148,66.0078406427285,56.6621327894133,64.427226260792,59.6411522580869,54.7853006620191,62.5791157289376,64.2432705505488,64.6637314187692,66.0977643647167,66.1309505076381,63.2605451217536,65.8227827462007,65.2274912077561,67.8282521078083,65.7658686316158,62.295434407344,57.8409797705986 ),  x=c( 2.0260852817446,9.03627189621329,3.7256769053638,10.362243656069,11.1665419801138,-1.36220900854096,5.39347683265805,10.4938666215166,5.72009020252153,4.39260629424825,11.3956668348983,4.34667818667367,7.48598889634013,6.01686762738973,-0.559054442681372,10.5975495856255,1.44522828096524,-1.41116653056815,2.59089006995782,11.3630510880612,10.4535504248925,7.69924768619239,6.96709539275616,11.9197768727317,7.1798811876215,7.91942655434832,5.61692434595898,6.31798828626052,2.04823632212356,0.0595910623669624 ), x2=c( 1,0,1,0,0,1,1,0,1,1,0,1,0,0,1,0,1,1,1,0,0,0,0,0,0,0,1,0,1,1 ))

View(challData)



```
sample(c(0,1), n, replace = TRUE)
if you type in console beginning of previous code and control arrow up, it shows list of other times you'v used it.
x2 is the choice of which law it will obey
original is the baseline model, the rest is the changes to the model are attached to a switch term. B2 next to on /off it allows you to change the intercept by beta2 the beta3 allows us to change the slope term. Nonsensical to square x2 because it's the switch

You can give data you don't use. You can put in switches you don't use. Extra x variables. Raise to the column power. can give as few points as possible, you could use transformations! Your goal is to become the weather and hide your secrets from all humanity

One of the arts is don't squeeze hard, don't go to far. Torture the data long enough and it'll confess to anything. You can start wwithout a quadratic (overfitting)

pairs(challData)
pairs(challData, col=as.factor(challData$x2))
1 pairs
2 view data
3 do lm summaries on opptions look at significance and rse
4 then plot wit h curve the options
5 can do diagnostic plot look at residuals
# these two lines are the same one is short version
lm21 <- lm(y~x*x2, data=challData)
lm21 <- lm(y ~ x + x2 + x:x2, challData)

plot(lm21)

