---
title: "midterm"
format: html
editor: visual
---

## library(mosaic)

library(tidyverse)

library(car)

\# 1 Perform a regression using the airquality data set in R where the response variable is Ozone and the explanatory variable is Wind.

\# What is the value of the MSE of this regression rounded to the nearest whole number?

View(airquality)

aqlm \<- lm(data=airquality, Ozone\~Wind)

summary(aqlm)

sum( aqlm\$res\^2 ) / 114

\# 2 Use the RailTrail data set in R from library(mosaic) to fit a regression to the following scatterplot. Compute the value of the residual for the solid dot labeled in the plot.

View(RailTrail)

RailTrail\[25,\]

rtlm \<- lm(hightemp \~ lowtemp, data=RailTrail)

summary(rtlm)

61-predict(rtlm, data.frame(lowtemp=35))

\# 3 accurate written equation

\# 4 What is the value of the MSE for the simple linear regression depicted in the plot?

sum(6.6\^2, 13.8\^2, 9.8\^2, 4.6\^2, 2\^2)/3

??# 5 For the Residuals vs Fitted plot shown, what is the value of for the solid orange dot?

\# 6 Suppose a researcher developed the theory that each one degree increase in daily average temperatures reduced average daily wind speed by 1/6 a mile per hour.

aq2lm \<- lm(data=airquality, Wind\~Temp)

summary(aq2lm)

calculate_t_value \<- function(estimate, std_error, number=0){

\# number is the value from hypotheses eg. H0:B0=3)

(estimate - number)/std_error

}

t_value \<- calculate_t_value(-0.17046, 0.02693, -0.167)

2\*pt(-abs(t_value), 151)

\# 7 Suppose the impossible happened and May 1st of a certain year ended up being 0 degrees Fahrenheit in New York City.

\# What daily average wind speed (in miles per hour) does your regression model expect for such a day?

?airquality

predict(aq2lm, data.frame(Temp=0))

??# 8 A regression is performed, and the following output obtained.

\# Use this output to create a 95% confidence interval for the true y-intercept

\# (B0) corresponding to this regression.

calculate_conf \<- function(estimate, min=0.025, max=0.975,

dof, std_error){

\# min and max for qt function)

estimate + qt(c(min, max), dof)\*std_error

}

calculate_conf(1.9845, dof=37, std_error=0.3521)

??# 9 Use the mtcars data set in R to perform a log(Y) transformation regression of mpg predicted by wt.

\# log transform is lamba of 0

lambda \<- 0

mtloglm \<- lm(log(mpg)\~wt, data=mtcars)

predict(mtloglm, data.frame(wt=2))

\## does this need retransformed?

e\^3.288344

exp(b.log\[1\] + b.log\[2\]\*x)

\# 10 see image

\# 11 definition

\# 12 kids feet regress girs width\~ length

\# Select the interval below that, according to your regression, we are 95% confident that it contains 95% of the actual widths of fourth-grade girl's feet that are 22 centimeters long.

?KidsFeet

GirlsFeet \<- filter(KidsFeet, sex == "G")

gflm \<- lm(data=GirlsFeet, width\~length)

predict(gflm, data.frame(length=22), interval = "confidence")

#13 The solid orange dot shown in the scatterplot below corresponds with which point on the residuals vs. fitted plot?

The Chrysler has a wt of 5.345 and a weight of 14.7 or nothing because I think the Chrysler pt shoud be near 15 on the fitted (x)

??#14 Consider the following scatterplot. Apply an appropriate Y transformation to the data shown in the plot.

\# State the R-squared value of the regression on the transformed data.

aqtrlm \<- lm(Ozone \~ Temp, data=airquality)

boxCox(aqtrlm) \# 0.25

lm(sqrt(sqrt(Y)) \~ X)

#15 Perform a linear regression that predicts the weight of chickens according to how old the chick is in days.

\# Which of the following correctly diagnoses the appropriateness of this regression?

?ChickWeight

cwlm \<- lm(data=ChickWeight, weight\~Time)

summary(cwlm)

windows(width = 10, height = 5)

par(mfrow=c(1,3))

plot(cwlm, which = 1:2) #1,3 & 2

plot(cwlm\$residuals) #5

```{r, fig.height=3, fig.width=3}
?ChickWeight
plot(weight ~ Time, data=ChickWeight)
cwlm <- lm(weight ~ Time, data=ChickWeight)
abline(cwlm)
plot(cwlm, which=2)
par(mfrow=c(1,3))
plot(cwlm, which = 1)
plot(cwlm$residuals)
```

```{r, fig.height=3, fig.width=3}
?Orange
plot(circumference ~ age, data=Orange)
olm <- lm(circumference ~ age, data=Orange)
abline(cwlm)
par(mfrow=c(1,3))
plot(olm, which=1)
plot(olm, which = 2)
plot(olm$residuals)
```

```{r, fig.height=3, fig.width=3}
plot(height ~ age, data=Loblolly)
lmlob <- lm(height ~ age, data=Loblolly)
abline(lmlob)
plot(lmlob, which=1)
```
