---
title: "Skills Quiz: Different Types of Models"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
    toc: true
    toc_float: true
---


## Instructions

Use this file to keep a record of your work as you complete the "Skills Quiz: Different Types of Models" assignment in Canvas.


```{r, message=FALSE, warning=FALSE}
library(mosaicData)
library(mosaic)
library(tidyverse)
```


----

<!-- Note: The {} after each Problem and Part header allows you to keep track of what work you have completed. Write something like {Done} once you complete each problem and your html file will then show you a nice summary of what you have "done" already. -->



## Problem 1 {}

Consider the scatterplot shown here of a single residence's monthly gas bill according to the month of the year. See `?Utilities` for more details on the data.

```{r, fig.height=3, fig.width=6}
?Utilities
plot(gasbill ~ month, data=Utilities, main="Single Residence in Minnesota", xlab="Month of the Year", ylab="Monthly Gas Bill (US Dollars)")
utlm <- lm(gasbill ~ month + I(month^2), data=Utilities)
summary(utlm)
b <- coef(utlm)
#curve(b[1] + b[2]*x + b[3]*x^2, add=TRUE)
curve(predict(utlm, newdata=data.frame(month=x)), add=TRUE, col="red", lwd=2)

par(mfrow=c(1,3))
plot(utlm, which = 1:2) #1,3 & 2
plot(utlm$residuals)
```



### Part (a)

Add the estimated regression function to the scatterplot above and state the function below.

<div class="YourAnswer">

$$
  \hat{Y}_i = 284.4558 - 76.6165 X_i + 5.4807 {X_i}^2

$$

</div>

### Part (b)

Diagnose the appropriateness of this regression model. How well does it fit the data? 

<span style="font-size:.8em;">  From the residuals vs. fitted values plot, the variance   
   to be constant and the model   
  meet the linearity requirement. Further, points 2, 65, and   
4
 are all possible outliers that should be checked for errors in encoding and possibly removed from the regression.

  From the Q-Q Plot of the residuals, normality  
   with the points going out of the "normality boundary lines" at the "norm quantiles" of   
-2
  on the left and   
1
  on the right of the graph.

 

  From the Residuals vs Order plot we see a possible pattern. Using View(Utilities) we see that the data is sorted by the   
year
 column (enter the name of the column the Utilities data is sorted by). It is reasonable to assume that the error terms    
   independent because of some consistent, though modest changes over time in the data.</span>

<div class="YourAnswer">

```{r}
par(mfrow=c(1,3))
plot(utlm, which = 1:2) #1,3 & 2
plot(utlm$residuals)
View(Utilities)
```

</div>


### Part (c)

What range of possible gas bill amounts do you predict for the September bill? How confident are you in your prediction?

<div class="YourAnswer">

```{r}
predict(utlm, data.frame(month=9), interval="prediction")
```

Type your answer here...

</div>


### Part (d)

Compute the mean of just September's gas bills and draw a horizontal reference line for this mean on your scatterplot. How does this mean compare to your prediction in Part (c)?

<div class="YourAnswer">

```{r}
sept <- Utilities %>%
  filter (month == "9")
mean(sept$gasbill)
38.84658-mean(sept$gasbill)
```

Type your answer here...

</div>




### Part (e)

State the values of $R^2$, $MSE$, and the residual standard error. What do each of these values tell us about the regression model?

<div class="YourAnswer">

```{r}
30.48^2 # square the RSE for the MSE
```
0.7937 r^2 The measure of the proportion of variability in gas bill prices explained by the regression model
103122.4 MSE The estimate of the model variance $\sigma^2$
30.48 RSE Gives an estimate of the standard deviation of the actual gas bill values around the model's average gas bill amount for each month


</div>

<br />




## Problem 2

View the `mtcars` dataset and corresponding help file `?mtcars`.

Perform a regression that predicts the miles per gallon `mpg` of the vehicle based on the quarter mile time `qsec` and transmission type `am` of the vehicle.


### Part (a)

Plot the data and your fitted regression model.

<div class="YourAnswer">

```{r}
#Type your code here...
```

</div>


### Part (b)

State the fitted regression model.

<div class="YourAnswer">

$$
 ...
$$

</div>

### Part (c)

Use `summary(...)` to perform an appropriate t test to determine if the interaction term is needed in this regression model.

<div class="YourAnswer">

```{r}
#Type your code here...
```

Type your answer here...

</div>


### Part (d)

Diagnose the appropriateness of this regression model. How well does it fit the data? 

<span style="font-size:.8em;">Be sure to provide diagnostic plots and supporting arguments for your claims.</span>

<div class="YourAnswer">

```{r}
#Type your code here...
```

Type your answer here...

</div>


### Part (e)

State the $R^2$ and residual standard error values for this model. What do these values show?

<div class="YourAnswer">

```{r}
#Type your code here...
```

Type your answer here...

</div>

<br />



## Problem 3

View the `mtcars` dataset and corresponding help file `?mtcars`.

Perform a regression that predicts the quarter mile time `qsec` of the vehicle based on the displacement of the engine `disp` and transmission type `am` of the vehicle according to the model:

$$
  \underbrace{Y_i}_\text{qsec} = \beta_0 + \beta_1 \underbrace{X_{1i}}_\text{disp} + \beta_2 X_{1i}^2 + \beta_3 \underbrace{X_{2i}}_\text{am == 1} + \beta_4 X_{1i}X_{2i} + \beta_5 X_{1i}^2X_{2i} + \epsilon_i
$$

where $\epsilon_i \sim N(0, \sigma^2)$.

### Part (a)

Plot the data and your fitted regression model.

<div class="YourAnswer">

```{r}
#Type your code here...
```

Type your answer here...

</div>

### Part (b)

State the fitted regression model.

<div class="YourAnswer">

$$
  \hat{Y}_i = ...
$$

</div>


### Part (c)

Use `summary(...)` to perform appropriate t tests to determine which interaction terms are needed in this regression model.

<div class="YourAnswer">

```{r}
#Type your code here...
```

</div>

### Part (d)

Diagnose the appropriateness of this regression model. How well does it fit the data? 

<span style="font-size:.8em;">Be sure to provide diagnostic plots and supporting arguments for your claims.</span>


<div class="YourAnswer">

```{r}
#Type your code here...
```

Type your answer here...

</div>

### Part (e) 

Drop the "least significant" term from the model. This is a silly statement really, but is an accepted practice when searching for a "best" model in regression analysis. What has changed?

<div class="YourAnswer">

```{r}
#Type your code here...
```

Type your answer here...

</div>










<style>

.YourAnswer {
  color: #317eac;
  padding: 10px;
  border-style: solid;
  border-width: 2px;
  border-color: skyblue4;
  border-radius: 5px;
}

</style>

 
 