---
title: "Skills Quiz: Different Types of Models"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
    toc: true
    toc_float: true
---


## Instructions

Use this file to keep a record of your work as you complete the "Skills Quiz: Different Types of Models" assignment in Canvas.


```{r, message=FALSE, warning=FALSE}
library(mosaicData)
library(mosaic)
library(tidyverse)
```


----

<!-- Note: The {} after each Problem and Part header allows you to keep track of what work you have completed. Write something like {Done} once you complete each problem and your html file will then show you a nice summary of what you have "done" already. -->



## Problem 1 {}

Consider the scatterplot shown here of a single residence's monthly gas bill according to the month of the year. See `?Utilities` for more details on the data.

```{r, fig.height=3, fig.width=6}
?Utilities
plot(gasbill ~ month, data=Utilities, main="Single Residence in Minnesota", xlab="Month of the Year", ylab="Monthly Gas Bill (US Dollars)")
utlm <- lm(gasbill ~ month + I(month^2), data=Utilities)
summary(utlm)
b <- coef(utlm)
#curve(b[1] + b[2]*x + b[3]*x^2, add=TRUE)
curve(predict(utlm, newdata=data.frame(month=x)), add=TRUE, col="red", lwd=2)

par(mfrow=c(1,3))
plot(utlm, which = 1:2) #1,3 & 2
plot(utlm$residuals)
```



### Part (a)

Add the estimated regression function to the scatterplot above and state the function below. Ans Quadratic seems best

<div class="YourAnswer">

$$
  \hat{Y}_i = 284.4558 - 76.6165 X_i + 5.4807 {X_i}^2

$$

</div>

### Part (b)

Diagnose the appropriateness of this regression model. How well does it fit the data? 

<span style="font-size:.8em;">  From the residuals vs. fitted values plot, the variance   
does not appear to be constant and the model should be considered to meet the linearity requirement. Further, points 2, 65, and 4 are all possible outliers that should be checked for errors in encoding and possibly removed from the regression.

  From the Q-Q Plot of the residuals, normality is clearly violated,
   with the points going out of the "normality boundary lines" at the "norm quantiles" of   -2 on the left and 1 on the right of the graph.


  From the Residuals vs Order plot we see a possible pattern. Using View(Utilities) we see that the data is sorted by the year column (enter the name of the column the Utilities data is sorted by). It is reasonable to assume that the error terms    
are not independent because of some consistent, though modest changes over time in the data.</span>

<div class="YourAnswer">

```{r}
par(mfrow=c(1,3))
plot(utlm, which = 1:2) #1,3 & 2
plot(utlm$residuals)
View(Utilities)
```

</div>


### Part (c)

What range of possible gas bill amounts do you predict for the September bill? How confident are you in your prediction? ans   At 95% confidence, our model shows that September's possible gas bill amounts range from   
0
 (because bill's can't be negative) to the upper limit with a predict of fit.

<div class="YourAnswer">

```{r}
predict(utlm, data.frame(month=9), interval="prediction")
```

Type your answer here...

</div>


### Part (d)

Compute the mean of just September's gas bills and draw a horizontal reference line for this mean on your scatterplot. How does this mean compare to your prediction in Part (c)?

<div class="YourAnswer">

```{r}
sept <- Utilities %>%
  filter (month == "9")
mean(sept$gasbill)
38.84658-mean(sept$gasbill) # 16.03158
```

So which one is right? Good question. While we haven't seen a lot of variability in past September bills for this data set, we do see high variability in many of the other months. It isn't unreasonable to assume that September could be prone to this same variability. The regression accounts for the variability of all months. Using just September isolates the data to just September's history, providing a more narrow view of the data. However, this particular regression model doesn't seem to perform very well on months 1, 9 and 10 (January, September and October) when it comes to capturing the average of those months. So... hard to say. But more work is needed to really capture the essence of this data.

</div>




### Part (e)

State the values of $R^2$, $MSE$, and the residual standard error. What do each of these values tell us about the regression model?

<div class="YourAnswer">

```{r}
30.48^2 # square the RSE for the MSE
```
0.7937 r^2 The measure of the proportion of variability in gas bill prices explained by the regression model
103122.4 MSE The estimate of the model variance $\sigma^2$
30.48 RSE Gives an estimate of the standard deviation of the actual gas bill values around the model's average gas bill amount for each month


</div>

<br />




## Problem 2

View the `mtcars` dataset and corresponding help file `?mtcars`.

Perform a regression that predicts the miles per gallon `mpg` of the vehicle based on the quarter mile time `qsec` and transmission type `am` of the vehicle.


### Part (a)

Plot the data and your fitted regression model.

<div class="YourAnswer">

```{r}
mt2lines.lm <- lm(mpg~qsec + am + qsec:am, data=mtcars)
b <- coef(mt2lines.lm)
# Then b will have 4 estimates:
# b[1] is the estimate of beta_0: -9.0099
# b[2] is the estimate of beta_1:  1.4385
# b[3] is the estimate of beta_2: -14.5107
# b[4] is the estimate of beta_3: 1.3214

ggplot(mtcars, aes(y=mpg, x=qsec, color=factor(am))) +
  geom_point(pch=21, bg="gray") +
  #geom_smooth(method="lm", se=F) + #easy way, but only draws the full interaction model. The manual way using stat_function (see below) is more involved, but more dynamic.
  stat_function(fun = function(x) b[1] + b[2]*x, color="skyblue") + #am==0 line
  stat_function(fun = function(x) (b[1]+b[3]) + (b[2]+b[4])*x, color="orange") + #am==1 line 
  scale_color_manual(name="Transmission (am)", values=c("skyblue","orange")) +
  labs(title="Two-lines Model using mtcars data set") 

summary(mt2lines.lm)
```

</div>


### Part (b)

State the fitted regression model.

<div class="YourAnswer">

$$
 \underbrace{Y_i}_\text{mpg} \underbrace{=}_{\sim}
\overbrace{\beta_0}^{\stackrel{\text{y-int}}{\text{baseline}}} +
\overbrace{\beta_1}^{\stackrel{\text{slope}}{\text{baseline}}}
\underbrace{X_{1i}}_\text{qsec} +
\overbrace{\beta_2}^{\stackrel{\text{change
in}}{\text{y-int}}}  \underbrace{X_{2i}}_\text{am} +
\overbrace{\beta_3}^{\stackrel{\text{change in}}{\text{slope}}}
\underbrace{X_{1i}X_{2i}}_\text{qsec:am} + \epsilon_i
$$

$$
 \underbrace{Y_i}_\text{mean mpg} =
-9.0099 +
1.4385
\underbrace{X_{1i}}_\text{qsec} +
-14.5107  \underbrace{X_{2i}}_\text{am==1} +
 1.3214
\underbrace{X_{1i}X_{2i}}_\text{qsec:am} + \epsilon_i
$$

</div>

### Part (c)

Use `summary(...)` to perform an appropriate t test to determine if the interaction term is needed in this regression model.

<div class="YourAnswer">

```{r}
0.07012
```

Type your answer here...

</div>


### Part (d)

Diagnose the appropriateness of this regression model. How well does it fit the data? 

<span style="font-size:.8em;">Be sure to provide diagnostic plots and supporting arguments for your claims.</span>

<div class="YourAnswer">

```{r, eval=FALSE, fig.height=4, fig.width=}
plot(mt2lines.lm, which=1:2)
plot(mt2lines.lm$residuals)
# okay for linearity, fine for constant variance, normality-a few outliers but otherwise okay, independence sure
```

Type your answer here...

</div>


### Part (e)

State the $R^2$ and residual standard error values for this model. What do these values show?

<div class="YourAnswer">

```{r}
# R2 0.6923 (adjusted R2 because of the number of params)
# R2 Measures the proportion of variability in the gas mileage of vehicles that is explained by the quarter mile time and transmission type of the vehicle.
# 3.343 Measures the variability (standard deviation) of gas mileage, in units of mpg, around the average gas mileage predicted for a vehicles with a certain quarter mile time and transmission type



#### part f
# Sometimes, when an interaction term is not significant, we may choose to use a reduced regression model that drops the interaction term. Mathematically this would turn the full two-lines model into this simpler mathematical model:
# same as above but no B3X1iX2i

mt2linesnoint.lm <- lm(mpg~qsec + am, data=mtcars)
b <- coef(mt2linesnoint.lm)

# When you drop the interaction term, the second half of the second line is no long the first slope plus the interaction slope (b[2] + b[4])*x, it's just the original slope b[2]*x
ggplot(mtcars, aes(y=mpg, x=qsec, color=factor(am))) +
  geom_point(pch=21, bg="gray") +
  stat_function(fun = function(x) b[1] + b[2]*x, color="skyblue") + #am==0 line
  stat_function(fun = function(x) (b[1]+b[3]) + b[2]*x, color="orange") + #am==1 line 
  scale_color_manual(name="Transmission (am)", values=c("skyblue","orange")) +
  labs(title="Two-lines Model using mtcars data set") 

summary(mt2linesnoint.lm)

```

Type your answer here...

</div>

<br />



## Problem 3

View the `mtcars` dataset and corresponding help file `?mtcars`.

Perform a regression that predicts the quarter mile time `qsec` of the vehicle based on the displacement of the engine `disp` and transmission type `am` of the vehicle according to the model:

$$
  \underbrace{Y_i}_\text{qsec} = \beta_0 + \beta_1 \underbrace{X_{1i}}_\text{disp} + \beta_2 X_{1i}^2 + \beta_3 \underbrace{X_{2i}}_\text{am == 1} + \beta_4 X_{1i}X_{2i} + \beta_5 X_{1i}^2X_{2i} + \epsilon_i
$$

where $\epsilon_i \sim N(0, \sigma^2)$.

### Part (a)

Plot the data and your fitted regression model.

<div class="YourAnswer">

```{r}

mt2qs.lm <- lm(qsec~disp +I(disp^2) + am + disp:am + I(disp^2):am, data=mtcars)
b <- coef(mt2qs.lm)
summary(mt2qs.lm)
format(6.607e-05) #can un scientific notation just move it positive or negative that number of spots  2.619e+01 is 26.9, -2.926e-03 is -0.002926, if it's e+00 it's just the number
# CHECK CLOSELY so you don't turn a positive with an e-xx to a negative number--only move the decimal

ggplot(mtcars, aes(x=disp, y=qsec, col=as.factor(am))) +
  geom_point() +
  stat_function(fun=function(x) b[1] + b[2]*x + b[3]*x^2, aes(color="0")) +
  stat_function(fun=function(x) (b[1] + b[4]) + (b[2] + b[5])*x + 
                  (b[3] + b[6])*x^2, aes(color="1")) +
  labs(title="The interaction of displacement andh transmisison type on quarter mile time", xlab="Displacement", ylab="Qsec") +
  scale_color_manual(name="Transmission (am)", values=c("skyblue","orange")) +
  theme_minimal()

 >Typically, we will drop the "least significant term" or term with the highest p-value first, then see if that changes the p-values of the other terms or not. Usually we will not drop all significant terms all at once. This is sometimes called a "backwards step-wise" model selection procedure.
```

Type your answer here...

</div>

### Part (b)

State the fitted regression model.

<div class="YourAnswer">

$$
  \hat{Y}_i = 26.19 - .04937X_1i + 0.0000661X_1i^2 -3.63X_2i - 0.0029263X_1iX_2i+.00001866X_1i^2 X_2i
$$

</div>


### Part (c)

Use `summary(...)` to perform appropriate t tests to determine which interaction terms are needed in this regression model.

<div class="YourAnswer">

```{r}
#Type your code here...the summary is in a It said to do it for B4 and B4. Then asked which could be removed the answer was either term not both (I guess because you have to drop one and check for the significance of the other before deciding)
```

</div>

### Part (d)

Diagnose the appropriateness of this regression model. How well does it fit the data? 

<span style="font-size:.8em;">Be sure to provide diagnostic plots and supporting arguments for your claims.</span>


<div class="YourAnswer">

```{r, eval=FALSE, fig.width=4, fig.height=4}
plot(mt2qs.lm, which=1:2)
plot(mt2qs.lm$residuals)
```

Type your answer here...

</div>

### Part (e) 

Drop the "least significant" term from the model. This is a silly statement really, but is an accepted practice when searching for a "best" model in regression analysis. What has changed?

<div class="YourAnswer">

```{r}
mt2qs2.lm <- lm(qsec~disp +I(disp^2) + am + I(disp^2):am, data=mtcars)
summary(mt2qs2.lm)
```

Type your answer here...

</div>

## Problem 4


```{r eval=FALSE}
# Simulate two quadratics
# y_i = b_0 + b_1X_i + b_2X_i^2 + b_3X_2 + b_4X_1X_2 + b_5X_i^2
library(ggplot2)
## Simulating Data from a Regression Model
## This R-chunk is meant to be played in your R Console.
## It allows you to explore how the various elements
## of the regression model combine together to "create"
## data and then use the data to "re-create" the line.

#Allows us to always get the same "random" sample
#Change to a new number to get a new sample
set.seed(111) 

#set the sample size
n <- 100 

# random uniform gives n random values in a uniform distribution from 15 to 45
#Gives n random values from a uniform distribution between 0 to 10.
X_i <- runif(n, -2, 2) 
X_2i <- sample(c(0,1), n, replace=TRUE) 

# Set   equation
beta0 <- -2 #Our choice for the y-intercept. 
beta1 <-  3 #Our choice for the slope. 
beta2 <-  4
beta3 <-  4
beta4 <-  2
beta5 <- -7
sigma <- .5 #Our choice for the std. deviation of the error terms.
#Gives n random values from a normal distribution with mean = 0, st. dev. = sigma.
epsilon_i <- rnorm(n, 0, sigma) 
    
#Create Y using the normal error quadratic regression model (don't forget beta3's X_2i)
Y_i <- beta0 + beta1*X_i + beta2*X_i^2 + beta3*X_2i + beta4*X_i*X_2i + beta5*(X_i^2)*X_2i + epsilon_i 
    
#Store the data as data
fabData <- data.frame(y=Y_i, x1=X_i, x2=X_2i) 
#View(fabData) 

quad2.lm <- lm(y ~ x1 + I(x1^2) + x2 + x1:x2 + I(x1^2):x2, data=fabData)
b <- coef(quad2.lm)
summary(quad2.lm)

  
ggplot(fabData, aes(y=y, x=x1, color=as.factor(x2))) + 
  geom_point(pch=16) +
  stat_function(fun = function(x) b[1] + b[2]*x + b[3]*x^2, lty=1, aes(color="Yhat1")) +
  stat_function(fun = function(x) (b[1] + b[4]) + (b[2] + b[5])*x + (b[3] + b[6])*(x^2), lty=1, aes(color="Yha2t")) +
  stat_function(fun = function(x) beta0 + beta1*x + beta2*x^2, lty=2, aes(color="E{Y}1")) +
  stat_function(fun = function(x) (beta0 + beta3) + (beta1 + beta4)*x + (beta2 + beta5)*(x^2), lty=2, aes(color="E{Y}2")) +
  labs(title="Quadratic Regression Relation Diagram") +
  scale_colour_manual("Legend", values = c("pink", "skyblue", "pink", "skyblue", "skyblue","pink")) +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  theme_minimal()

# Could pass in the x2 and set it to 0 the first time, 1 the second
# stat_function(fun = function(x, x2=0) b[1] + b[2]*x + b[3]*x^2 + b[4]*x2 + b[5]*x*x2 + b[6]*x^2*x2, lty=1, aes(color="Yhat1")) +
  






```

## Further Exploration

```{r}
set.seed(101) #Allows us to always get the same "random" sample                
#Change to a new number to get a new sample  

n <- 100 #30 #set the sample size, 30 was original  

X_i <- runif(n, -5, 55)     #Gives n random values from a uniform distribution between 15 to 45.  

X_2i <- sample(c(0,1,2,4,10), n, replace=TRUE) # round 3
#sample(c(0,1,2), n, replace=TRUE) #round 2 code
#sample(c(0,1), n, replace=TRUE)      #Gives n 0's and 1's randomly    this was the original code

 

beta0 <- 3 #Our choice for the y-intercept.  

beta1 <- 2.5 #Our choice for the slope.    

beta2 <- -20 #Our choice for the change in y-intercept from the first line to the second line.

beta3 <- 3.5 #Our choice for the change in slope from the first line to the second line.     

 

sigma <- 3 #Our choice for the std. deviation of the error terms.  

epsilon_i <- rnorm(n, 0, sigma)     #Gives n random values from a normal distribution with mean = 0, st. dev. = sigma.  

 

Y_i <- beta0 + beta1*X_i + beta2*X_2i + beta3*X_i*X_2i + epsilon_i     #Create Y using the Two-Lines regression model  

 

fabData <- data.frame(y=Y_i, x=X_i, x2=X_2i)     #Store the data as data  

#In the real world, we begin with data (like fabData) and try to recover the model that  

# (we assume) was used to created it.  

 

fab.lm <- lm(y ~ x + x2 + x:x2, data=fabData) #Fit an estimated two-lines regression model to the fabData.  

summary(fab.lm) #Summarize your model.  

 

plot(y ~ x, data=fabData, col=as.factor(x2), pch=16) #Plot the data.  

b <- coef(fab.lm)   #Add the estimated regression line to your plot.  

x2 = 0  #draw the x2=0 baseline line

curve(b[1] + b[2]*x + b[3]*x2 + b[4]*x*x2, add=TRUE)  

x2 = 1   #draw the x2=1 line (changed line)

curve(b[1] + b[2]*x + b[3]*x2 + b[4]*x*x2, add=TRUE, col="red")

x2=2

  curve(beta0 + beta1*x + beta2*x2 +beta3*x*x2, add=TRUE, lty=2, col="green")
  
x2=4

  curve(beta0 + beta1*x + beta2*x2 +beta3*x*x2, add=TRUE, lty=2, col="cornflowerblue")
  
x2=10

  curve(beta0 + beta1*x + beta2*x2 +beta3*x*x2, add=TRUE, lty=2, col="turquoise")
```

## four line model

```{r}
set.seed(101) #Allows us to always get the same "random" sample                
#Change to a new number to get a new sample  

n <- 100 #30 #set the sample size, 30 was original  

X_i <- runif(n, -5, 55)     #Gives n random values from a uniform distribution between 15 to 45.  

X_2i <- sample(c(0,1), n, replace=TRUE) 

X_3i <- sample(c(0,1), n, replace=TRUE) #info said 0 or 1

beta0 <- 3 #Our choice for the y-intercept.  

beta1 <- 2.5 #Our choice for the slope.    

beta2 <- -20 #Our choice for the change in y-intercept from the first line to the second line.

beta3 <- 3.5 #Our choice for the change in slope from the first line to the second line.     

beta4 <- 1.5
beta5 <- 2
beta6 <- -1

sigma <- 3 #Our choice for the std. deviation of the error terms.  

epsilon_i <- rnorm(n, 0, sigma)     #Gives n random values from a normal distribution with mean = 0, st. dev. = sigma.  

 

Y_i <- beta0 + beta1*X_i + beta2*X_2i + beta3*X_i*X_2i + beta4*X_3i + beta5*X_i*X_3i + beta6*X_i*X_2i*X_3i + epsilon_i     #Create Y using the Two-Lines regression model  

 

fabData <- data.frame(y=Y_i, x=X_i, x2=X_2i)     #Store the data as data  

#In the real world, we begin with data (like fabData) and try to recover the model that  

# (we assume) was used to created it.  

 

fab.lm <- lm(y ~ x + x2 + x:x2, data=fabData) #Fit an estimated two-lines regression model to the fabData.  

summary(fab.lm) #Summarize your model.  

 

plot(y ~ x, data=fabData, col=as.factor(x2), pch=16) #Plot the data.  

b <- coef(fab.lm)   #Add the estimated regression line to your plot.  

x2 = 0  #draw the x2=0 baseline line
X3 = 0 

curve(b[1] + b[2]*x + b[3]*x2 + b[4]*x*x2, add=TRUE)  

x2 = 1   #draw the x2=1 line (changed line)
X3 = 0 

curve(beta0 + beta1*x + beta2*x2 + beta3*x*x2 + beta4*x3 + beta5*x*x3 + beta6*x*x2*x3, add=TRUE, lty=2, col="red")

x2 = 0   
X3 = 1

curve(beta0 + beta1*x + beta2*x2 + beta3*x*x2 + beta4*x3 + beta5*x*x3 + beta6*x*x2*x3, add=TRUE, lty=2, col="green")
  
x2 = 1   
X3 = 1

curve(beta0 + beta1*x + beta2*x2 + beta3*x*x2 + beta4*x3 + beta5*x*x3 + beta6*x*x2*x3, add=TRUE, lty=2, col="cornflowerblue")

```

<style>

.YourAnswer {
  color: #317eac;
  padding: 10px;
  border-style: solid;
  border-width: 2px;
  border-color: skyblue4;
  border-radius: 5px;
}

</style>

 
 